---
title: "Metadata_Harmonization_New"
output: html_document
---

```{r}
# Load libraries
library(tidyverse)
library(stringr)
library(dplyr)
library(here)
```

```{r}
# Read in combined metadata
metadata_orig <- readRDS(here("output_data/non_mouse_data/updated_data/combined_species_metadata_raw.rds"))

# Check
head(metadata_orig)
```

# Check & correct mammal names

```{r}
# Check species
unique(metadata_orig$organism_scientific_name)

```

```{r}
# Update species names to only have 2 names
metadata_orig <- metadata_orig %>%
  mutate(organism_scientific_name = recode(organism_scientific_name,
                            "Enhydra lutris kenyoni" = "Enhydra lutris",
                            "Mustela putorius furo" = "Mustela putorius",
                            "Sus scrofa domesticus" = "Sus scrofa"))

# Check if mammals are renamed
unique(metadata_orig$organism_scientific_name)
```

# Remove duplicate phrases and repeated words function 

```{r}
remove_duplicates <- function(s) {
  # Leave NAs as is
  if (is.na(s)) return(NA_character_)

  # Remove any extra white space
  s <- trimws(gsub("\\s+", " ", s))
  
  # Make empty strings R's NA val
  if (s == "") return(NA_character_)

  # Split the string into words
  words <- strsplit(s, " ", fixed = TRUE)[[1]]
  lower <- tolower(words) # convert to lowercase
  n <- length(words) # number of words

  # Check for repeated phrases
  for (k in seq_len(floor(n / 2))) {
    if (n %% k == 0) {
      phrase <- lower[1:k] # 1st phrase
      repeated <- rep(phrase, n / k) # repeat the phrase

      # Check if the string is the repeated phrase
      if (all(lower == repeated)) {
        return(paste(words[1:k], collapse = " "))
      }
    }
  }

  # If there are duplicated words, remove them (ex. PAIRED paired Paired: only keep 1)
  words <- words[c(TRUE, lower[-1] != lower[-n])]

  # Put the words back into the string
  paste(words, collapse = " ")
}
```

# First_nonempty function to get the first value across multiple columns to put in a harmonized column

```{r}
# Function to retrieve the first non-empty (no NAs) entry from the cols given
first_nonempty <- function(...) {                
  x <- c(...) # combine all inputs into one vector
  x <- trimws(as.character(x)) # remove any extra white space
  x[x %in% c("", "NA", "NaN", "NULL", "missing", "Missing", "MISSING", "unknown", "Unknown", "UNKNOWN", "not applicable", "Not applicable", "none")] <- NA_character_ # replace NaN's with R's NA val
  x <- x[!is.na(x)] # remove NAs
  if (length(x) == 0) NA_character_ else x[[1]] # return NA if empty, otherwise first value
}

```

# Function to replace common missing data terms with NA

```{r}
clean_df_missing <- function(df) {
  # Missing data terms to be replaced with R's NA
  missing_terms <- c("na", "nan", "null", "missing", "unknown", "not applicable", "none", "")
  
  # Iterate through the df cols
  for (col in names(df)) {
    # Only check character cols
    if (is.character(df[[col]])) {
      # Remove any extra white space and make lowercase to compare to missing_terms
      x_clean <- trimws(tolower(df[[col]]))
      
      # Replace missing terms with NA
      df[[col]][x_clean %in% missing_terms] <- NA
    }
  }
  
  return(df)
}

```

# Remove NaNs, duplicate phrases and words and replace missing data with NA

```{r}
metadata_harmonized <- metadata_orig

# Remove the string "nan" from the df
metadata_harmonized[] <- lapply(metadata_harmonized, function(x) {
  if (is.character(x)) {
    x <- gsub("\\bnan\\b", "", x, ignore.case = TRUE) # remove nan
    x <- trimws(gsub("\\s+", " ", x)) # remove extra white space
    x[x == ""] <- NA_character_ # make empty strings R's NA val 
  }
  x # return the column
})

```

```{r}
# Apply the remove_duplicates() function to the df to remove duplicate phrases and words
metadata_harmonized[] <- lapply(metadata_harmonized, function(x) {
  if (is.character(x)) {
    # Remove repeated phrases in each cell and return the cleaned col
    vapply(x, remove_duplicates, character(1))
  } else {
    x # return cols that aren't characters unchanged
  }
})

```

```{r}
# Apply the clean_df_missing() function to the df to put NAs for any common missing terms
metadata_harmonized <- clean_df_missing(metadata_harmonized)

```

# Harmonize library layout column & rename single-cell/bulk column

```{r}
# Check the library layout column
unique(metadata_harmonized$library_layout)
```

```{r}
# Make all values uppercase in library layout
metadata_harmonized$library_layout <- toupper(metadata_harmonized$library_layout)

# Check
unique(metadata_harmonized$library_layout)

```

```{r}
# Merge single-cell/bulk and single_bulk 
metadata_harmonized <- metadata_harmonized %>%
  mutate(single_bulk = coalesce(single_bulk, `single-cell/bulk`)) %>%
  select(-`single-cell/bulk`) # drop this col

# Check
unique(metadata_harmonized$single_bulk)
```

# Harmonize biological sex

```{r}
# Check cols that have sex
sex_cols <- grep("sex|gender|reproductive", names(metadata_harmonized), ignore.case = TRUE, value = TRUE)

sex_cols
```

```{r}
# Check for more cols
metadata_harmonized %>%
  select(where(is.character)) %>%
  select(where(~ any(grepl("male|female", ., ignore.case = TRUE)))) %>%
  names()

```

```{r}
# Add more cols to sex_cols
sex_cols <- c(sex_cols, "description", "animal")

# Check unique values of the sex cols
unique(unlist(metadata_harmonized[sex_cols]))
```

```{r}
# Create a temp combined col for all sex col values using first_nonempty()
metadata_harmonized$sex_combined <- apply(metadata_harmonized[, sex_cols], 1, first_nonempty)

# Create paipu_sex col for harmonized values
metadata_harmonized$paipu_sex <- "UNKNOWN"

# Harmonize biological sex
# Female
metadata_harmonized$paipu_sex[
  grepl("\\bfemale\\b", metadata_harmonized$sex_combined, ignore.case = TRUE) |
  metadata_harmonized$sex_combined %in% c("F", "FS")
] <- "FEMALE"

# Male
metadata_harmonized$paipu_sex[
  grepl("\\bmale\\b", metadata_harmonized$sex_combined, ignore.case = TRUE) |
  metadata_harmonized$sex_combined %in% c("MN")
] <- "MALE"

# Drop the temp col sex_combined
metadata_harmonized <- metadata_harmonized %>% select(-sex_combined)
```

```{r}
# Check
table(metadata_harmonized$paipu_sex)
```

# Harmonize reproductive status

```{r}
# Check cols that have reproductive info
reproductive_cols <- metadata_harmonized %>%
  select(where(is.character)) %>%
  select(where(~ any(grepl("neuter|castrate|intact|spay", ., ignore.case = TRUE)))) %>%
  names()

reproductive_cols
```

```{r}
# Check unique vals
unique(unlist(metadata_harmonized[reproductive_cols]))
```

```{r}
# Create a temp combined col for all reproductive col values using first_nonempty()
metadata_harmonized$reproductive_combined <- apply(metadata_harmonized[, reproductive_cols], 1, first_nonempty)

# Create paipu_reproductive_status col for harmonized values
metadata_harmonized$paipu_reproductive_status <- NA_character_

# Neutered
metadata_harmonized$paipu_reproductive_status[
  grepl("neuter|spay|castrate", metadata_harmonized$reproductive_combined, ignore.case = TRUE) |
  metadata_harmonized$reproductive_combined %in% c("FS", "MN")
] <- "NEUTERED"

# Intact
metadata_harmonized$paipu_reproductive_status[
  grepl("\\bintact\\b", metadata_harmonized$reproductive_combined, ignore.case = TRUE)
] <- "INTACT"

# Drop the temp col sex_combined
metadata_harmonized <- metadata_harmonized %>% select(-reproductive_combined)
```


# Harmonize platform_instrument_model

```{r}
# Check unique values
unique(metadata_harmonized$platform_instrument_model)

```

```{r}
# Replace all NA with Unknown
metadata_harmonized$platform_instrument_model[is.na(metadata_harmonized$platform_instrument_model)] <- "Unknown"

# Check
unique(metadata_harmonized$platform_instrument_model)
```

# Harmonize tumor grade

```{r}
# Check cols that have tumor grade
grep("grad", names(metadata_harmonized), ignore.case = TRUE, value = TRUE)

# Check unique vals
unique(metadata_harmonized$tumor_grade)
unique(metadata_harmonized$tumor_grading)
```

```{r}
# Apply first_nonempty() function to create a harmonized tumor grade col
metadata_harmonized$paipu_tumor_grade <- mapply(
  first_nonempty,
  metadata_harmonized$tumor_grade,
  metadata_harmonized$tumor_grading
)

# Check
unique(metadata_harmonized$paipu_tumor_grade)
```

# Harmonize dosage

```{r}
# Check cols that have dosage info
grep("dos|concentr", names(metadata_harmonized), ignore.case = TRUE, value = TRUE)

```

```{r}
# Check unique vals
unique(metadata_harmonized$dose)
unique(metadata_harmonized$concentration)
unique(metadata_harmonized$dosage)
```

```{r}
# Apply first_nonempty() function to create a harmonized dosage col
metadata_harmonized$paipu_dosage <- mapply(
  first_nonempty,
  metadata_harmonized$dose,
  metadata_harmonized$concentration,
  metadata_harmonized$dosage
)

# Check
unique(metadata_harmonized$paipu_dosage)
```

# Harmonize infection

```{r}
# Check cols that have infection info
grep("infect", names(metadata_harmonized), ignore.case = TRUE, value = TRUE)

```

```{r}
# Function to combine two infection columns
combine_infections <- function(x, y) {
  # Put both cols into 1
  vals <- c(x, y)
  vals <- vals[!is.na(vals)] # remove NAs
  
  # If vals doesn't have anything, return NA
  if (length(vals) == 0) {
    NA
  } else { # Otherwise, combine the 2 cols with a ';' separator
    paste(unique(vals), collapse = "; ") # unique avoids duplication
  }
}

# Combine infection_1 and infection_2 into 1 col before harmonizing
metadata_harmonized$infection_combined <- mapply(
  combine_infections,
  metadata_harmonized$infection_1,
  metadata_harmonized$infection_2
)
```

```{r}
# Apply first_nonempty() function to create a harmonized infection col
metadata_harmonized$paipu_infection <- mapply(
  first_nonempty,
  metadata_harmonized$infection,
  metadata_harmonized$infect,
  metadata_harmonized$infection_combined,
  metadata_harmonized$fcov_infection
)

# Drop temp col infection_combined
metadata_harmonized <- metadata_harmonized %>%
  select(-infection_combined)

# Check
unique(metadata_harmonized$paipu_infection)
```

# Harmonize breed

```{r}
# Check cols that have dosage info
grep("breed", names(metadata_harmonized), ignore.case = TRUE, value = TRUE)
```


```{r}
# Apply first_nonempty() function to create a harmonized breed col
metadata_harmonized$paipu_breed <- mapply(
  first_nonempty,
  metadata_harmonized$breed,
  metadata_harmonized$dog_breed
)

# Check
unique(metadata_harmonized$paipu_breed)
```

# Harmonize cell type

```{r}
# Find possible columns that relate to cell type
grep("cell.?type|celltype|cel.?type|celle.?type|cell.?typ", names(metadata_harmonized), ignore.case = TRUE, value = TRUE)

```

```{r}
# List of cell type cols
celltype_cols <- c("cell_type", "cell_type.1", "cell_type.2", "cell_type.3", "injected_cell_type", "tissue_cell_type_source")

# Apply first_nonempty() function to create a harmonized cell type col
metadata_harmonized$paipu_cell_type <- do.call(
  function(...) {
    # Pass all cell type cols into first_nonempty() for each row
    mapply(first_nonempty, ..., USE.NAMES = FALSE)},
  metadata_harmonized[celltype_cols] # list of cell type cols to use
)

# Create a flag where more than one cell type col has data
celltype_candidates <- metadata_harmonized[celltype_cols] # subset cell type cols

# Count how many cell type cols are filled per row
celltype_filled_count <- rowSums(
  !is.na(celltype_candidates) & trimws(as.matrix(celltype_candidates)) != ""
)

metadata_harmonized$paipu_cell_type_multiple_sources <- celltype_filled_count > 1 # true if more than one col had a cell type

# Check how many rows have more than 1 source (38)
table(metadata_harmonized$paipu_cell_type_multiple_sources) 

```

# Harmonize tissue
```{r}
# Check cols that mention tissue or organ
grep("tissue|organ|organism_part|organism|tissuetype|tissue_type|organ_part|tissue_of|source_tissue|tissue_origin", names(metadata_harmonized), ignore.case = TRUE, value = TRUE)
```


```{r}
# List of tissue cols
tissue_cols <- c(
  "tissue",
  "organ",
  "organism_part",
  "tissuetype",
  "tissue_type",
  "tissue_source"
)
```

```{r}
# Apply first_nonempty() function to create a harmonized tissue col
metadata_harmonized$paipu_tissue <- do.call(
  # Pass all cell type cols into first_nonempty() for each row
  function(...) mapply(first_nonempty, ..., USE.NAMES = FALSE),
  metadata_harmonized[tissue_cols] # list of tissue cols to use
)

# Create a flag where more than one tissue col has data
tissue_candidates <- metadata_harmonized[tissue_cols]

# Count how many tissue cols are filled per row
tissue_filled_count <- rowSums(
  !is.na(tissue_candidates) & trimws(as.matrix(tissue_candidates)) != ""
)

metadata_harmonized$paipu_tissue_multiple_sources <- tissue_filled_count > 1 # true if more than one col had a cell type

# Check how many rows have more than 1 source (128)
table(metadata_harmonized$paipu_tissue_multiple_sources)
```

```{r}
length(unique(metadata_harmonized$paipu_tissue))
```

```{r}
# Helpers for regex
sep <- "[-_\\s]*" # any dash, underscore or spaces between words
edge <- "(^|[^A-Za-z0-9])" # start of string (only words separated from others and not embedded inside)
tail <- "($|[^A-Za-z0-9])" # end of string  (only words separated from others and not embedded inside)

```

```{r}
# Function to create a regex that matches a whole word
word_match <- function(word) {
  paste0(edge, word, tail)
}

# Function to create a regex that matches a phrase with separators between words
phrase_match <- function(words) {
  paste0(edge, paste(words, collapse = sep), tail)
}

# Function to combine many regex patterns with 'or'
any_match <- function(...) {
  parts <- unlist(list(...))
  paste0("(", paste(parts, collapse = "|"), ")")
}

```

```{r}
tissue_patterns <- c(
  bone_marrow = any_match(phrase_match(c("bone","marrow")), 
      word_match("bm")),
  
  lymph_node = any_match(
      phrase_match(c("lymph","node")),
      phrase_match(c("lymph","nodes")),
      word_match("tdln"),
      word_match("dln"),
      phrase_match(c("tumor","draining","lymph","node"))),
  
  spleen = any_match(word_match("spleen"), 
      word_match("splenic")),
  
  blood = any_match(word_match("blood"),
      phrase_match(c("peripheral","blood"))),
  
  serum = word_match("serum"),

  # nervous system
  spinal_cord = phrase_match(c("spinal","cord")),
  
  brain = any_match(
      word_match("brain"),
      word_match("neocortex"),
      phrase_match(c("prefrontal","cortex")),
      phrase_match(c("brain","cortex")),
      phrase_match(c("cerebral","cortex")),
      word_match("cerebellum"), word_match("cerebella"),
      word_match("hippocampus"),
      word_match("thalamus"),
      word_match("brainstem"),
      phrase_match(c("olfactory","bulb"))),

  # major organs
  lung = any_match(word_match("lung"),
      word_match("lungs")),
  
  liver = any_match(word_match("liver"), 
      word_match("hepatic")),
  
  pancreas = any_match(word_match("pancreas"), 
      word_match("pancreata"),
      word_match("pancreatic")),
  
  kidney = any_match(word_match("kidney"),
      word_match("renal"),
      phrase_match(c("renal","cortex"))),
  
  bladder = any_match(word_match("bladder"),
      word_match("urothelial"),
      phrase_match(c("urinary","bladder"))),
  
  heart = word_match("heart"),
  
  aorta = word_match("aorta"),

  # gastrointestinal
  esophagus = any_match(word_match("esophagus"),
      word_match("esophageal")),
  
  oral = any_match( word_match("oral"),
      phrase_match(c("oral","mucosa")),
      phrase_match(c("oral","cavity")),
      phrase_match(c("oral","tongue")),
      phrase_match(c("oral","buccal")),
      word_match("gingiva"),
      word_match("buccal")),
  
  stomach   = any_match(word_match("stomach"), 
    word_match("gastric"),
    word_match("forestomach")),
  
  small_intestine = any_match(
    phrase_match(c("small","intestine")),
    phrase_match(c("small","intestinal")),
    word_match("ileum"),
    word_match("jejunum"),
    word_match("jejunal"),
    word_match("duodenum")),
  
  large_intestine = any_match(
    word_match("colon"),
    word_match("colonic"),
    word_match("colorectal"),
    word_match("cecal"),
    word_match("caecum"),
    word_match("rectal")),

  # skin
  skin = any_match(word_match("skin"),
    word_match("epidermis"), 
    word_match("epidermal")),
  
  # muscle
  muscle = any_match(
    word_match("muscle"),
    phrase_match(c("skeletal","muscle")),
    word_match("gastrocnemius"),
    word_match("soleus"),
    phrase_match(c("tibialis","anterior")),
    word_match("tibialis"),
    phrase_match(c("ta","muscle"))),

  # reproductive
  mammary = any_match(word_match("mammary"), 
    word_match("breast")),
  
  ovary = any_match(word_match("ovary"),
    word_match("ovarian"), 
    word_match("oviduct")),
  
  uterus = any_match(word_match("uterus"), 
    word_match("uterine")),
  
  testis = any_match(word_match("testis"),
    word_match("testes"),
    word_match("scrotum"),
    phrase_match(c("seminal","vesicle"))),
  
  prostate = any_match(word_match("prostate"),
    word_match("prostatic")),

  # endocrine glands
  adrenal_gland = any_match(phrase_match(c("adrenal","gland")),
    word_match("adrenal")),
  
  thyroid = word_match("thyroid"),
  
  pituitary = word_match("pituitary"),
  
  salivary_gland = any_match(
    phrase_match(c("salivary","gland")),
    phrase_match(c("submandibular","gland")),
    word_match("smg")),

  # other tissues
  omentum = any_match(word_match("omentum"), 
    word_match("omental")),
  
  peritoneum = any_match(word_match("peritoneum"),  
    phrase_match(c("peritoneal","cavity"))),
    diaphragm = word_match("diaphragm"),

  # eyes
  eye = any_match(
    word_match("eye"),
    word_match("ocular"),
    word_match("retina"),
    phrase_match(c("optic","nerve")))
)
```


```{r}
# Function to fix misspellings in tissue col
fix_misspellings <- function(s) {
  s <- str_replace_all(s, "\\brecal\\b", "rectal")
  s # return correct spelling
}

# Function to map paipu_tissue to a specific tissue
extract_explicit_tissue <- function(x) {
  if (is.na(x)) return(NA_character_) # if NA, return NA
  
  s <- str_to_lower(trimws(x)) # make lowercase and remove extra white space
  
  s <- fix_misspellings(s)
  
  if (s == "") return(NA_character_) # if empty, return NA
  
  if (s == "cortex") return("brain") # all 'cortex' are brain related in the metadata (checked manually)
  if (str_detect(s, phrase_match(c("renal","cortex")))) return("kidney") # map renal cortex explicitly to kidney so it isn't mapped to brain
  
  # Iterate through tissues
  for (tissue in names(tissue_patterns)) {
    if (str_detect(s, tissue_patterns[[tissue]])) { # check if a tissue pattern matches the string
      return(tissue) # return the first matched tissue
    }
  }
  NA_character_ # return NA if no tissue matched
}

# Label tissues explicitly
metadata_harmonized$paipu_tissue_explicit <- vapply(
  metadata_harmonized$paipu_tissue,
  extract_explicit_tissue,
  FUN.VALUE = character(1) # one string per row
)
```

```{r}

# Function to detect terms that could mess up tissue inference 
has_non_tissue_terms <- function(s) {
  if (is.na(s) || s == "") return(FALSE) # NA and empty wouldn't have non-tissue terms  
  
  str_detect(s, paste(
    c(
    # metastasis and secondary terms
    "metastas", "\\bmet\\b", "secondary", "disseminat", "recurr",
    # implantation, model and location terms
    "subcut", "orthotop", "intracran", "intraderm", "\\bflank\\b",
    "xenograft", "allograft", "syngen", "\\bgraft\\b", "implan", "inoculat",
    # cell type terms
    "cell\\s*line", "cellline", "organoid", "spheroid", "\\b3d\\b",
    "sorted", "\\bcd\\d+\\+?", "cd45\\+", "\\btils?\\b", "\\bpbmc\\b",
    "macrophage", "\\bt\\s*cells?\\b", "\\bb\\s*cells?\\b", "endothelial",
    "fibroblast", "immune\\s*cells?", "splenocytes", "single\\s*cell", "circulating",
    # fluids and non-tissue terms
    "\\bserum\\b", "ascites", "pleural", "effusion", "lavage", "faeces|fecal"
    ),
    collapse = "|"
  ))
}

# Function to map cancer type terms to an inferred tissue
get_tissue_from_cancer_type <- function(s) {
  if (is.na(s) || s == "") return(NA_character_) # NA and empty wouldn't have matches
  
  # Brain
  if (str_detect(s, "\\bglioblastoma\\b|\\bgbm\\b|\\bglioma\\b|\\boligodendroglioma\\b|\\bmedulloblastoma\\b|\\bdipg\\b")) return("brain")
  
  # Skin
  if (str_detect(s, "\\bmelanoma\\b|\\buv(e)?al\\s+melanoma\\b")) return("skin")
  
  # Pancreas
  if (str_detect(s, "\\bpdac\\b|pancreatic\\s+ductal\\s+adenocarcinoma")) return("pancreas")
  
  # Liver
  if (str_detect(s, "hepatocellular\\s+carcinoma|\\bhcc\\b|hepatoblastoma")) return("liver")
  
  # Lung
  if (str_detect(s, "\\bnsclc\\b|\\bsclc\\b|lung\\s+adenocarcinoma|lung\\s+squamous")) return("lung")
  
  # Colon
  if (str_detect(s, "\\bcrc\\b|colorectal|colon\\s+adenocarcinoma")) return("large_intestine")
  
  # Bladder
  if (str_detect(s, "urothelial|bladder\\s+cancer")) return("bladder")
  
  # Kidney
  if (str_detect(s, "\\brcc\\b|renal\\s+cell\\s+carcinoma")) return("kidney")
  
  # Prostate
  if (str_detect(s, "prostate\\s+cancer")) return("prostate")
  
  # Ovary
  if (str_detect(s, "ovarian\\s+cancer|high\\s+grade\\s+serous")) return("ovary")
  
  # Breast
  if (str_detect(s, "\\btnbc\\b|triple\\s+negative|breast\\s+cancer|mammary\\s+carcinoma")) return("mammary")
  
  NA_character_ # otherwise, NA
}

# Normalize tissue strings
tissue_strings <- metadata_harmonized$paipu_tissue |>
  str_to_lower() |>
  str_replace_all("[^a-z0-9]", " ") |> # remove punctuation
  str_squish()

# Only infer tissue if no explicit tissue is there
metadata_harmonized$paipu_tissue_inferred <- vapply(
  seq_along(tissue_strings),
  function(i) {
    s <- tissue_strings[i]
    # Skip rows that have explicit tissue already
    if (!is.na(metadata_harmonized$paipu_tissue_explicit[i])){
      return (NA_character_)
    }
    
    # Skip rows that have non-tissue terms
    if (has_non_tissue_terms(s)) {
      return(NA_character_)
    }
    
    # Infer tissue from cancer type terms
    get_tissue_from_cancer_type(s)
  },
  FUN.VALUE = character(1)
)
```

```{r}
# Determine final tissue label by choosing the explicit tissue if present, otherwise use inferred tissue label
metadata_harmonized$paipu_tissue_final <- ifelse(
  !is.na(metadata_harmonized$paipu_tissue_explicit),
  metadata_harmonized$paipu_tissue_explicit,
  metadata_harmonized$paipu_tissue_inferred
)

# Clean the final tissue label to replace underscores with spaces
metadata_harmonized$paipu_tissue_final <- str_replace_all(metadata_harmonized$paipu_tissue_final,"_", " ")
```

```{r}
length(unique(metadata_harmonized$paipu_tissue_final))
unique(metadata_harmonized$paipu_tissue_final)
```

```{r}
# Set NAs in paipu_tissue_final to 'unlabeled'
metadata_harmonized$paipu_tissue_final[is.na(metadata_harmonized$paipu_tissue_final)] <- "unlabeled"


# Check
table(metadata_harmonized$paipu_tissue_final, useNA = "ifany")
```

```{r}
# Read in manual tissue labels
tissue_labels <- read.csv(here("output_data/non_mouse_data/after_cancer_harmonization/manual_tissues.csv"))

# Label empty manual labels as NA
tissue_labels <- tissue_labels %>%
  mutate(manual_tissue = na_if(manual_tissue, ""))

head(tissue_labels)
```

```{r}
# Replace final tissue labels with manual labels if it exists
metadata_harmonized <- metadata_harmonized %>%
  # Add manual tissue labels to the df using run_accession
  left_join(tissue_labels %>% select(run_accession, manual_tissue), by = "run_accession") %>%
  
  # Replace paipu_tissue_final with manual_tissue where available
  mutate(paipu_tissue_final = coalesce(manual_tissue, paipu_tissue_final)) %>%
  
  # Remove manual_tissue col
  select(-manual_tissue)

```

```{r}
# Check
table(metadata_harmonized$paipu_tissue_final)
```


# Harmonize cancer type
## Infer cancer type from tissue (for the generic cancer terms - cancer*, tumor*, onco*)

```{r}
table(metadata_harmonized$cancer_type)
```


```{r}
# Flag generic cancer terms
is_generic_cancer <- function(x) {
  !is.na(x) & str_detect(x, "^(cancer\\*|tumor\\*|onco\\*)$")
}

```

```{r}
# If cancer_type is generic and paipu_tissue_final is not 'unlabeled', infer cancer type from tissue label
metadata_harmonized$paipu_cancer_type <- ifelse(
  is_generic_cancer(metadata_harmonized$cancer_type) &
    metadata_harmonized$paipu_tissue_final != "unlabeled",

  # Make the cancer type be tissue + "cancer"
  paste(metadata_harmonized$paipu_tissue_final, "cancer"),

  # Otherwise stay the same
  metadata_harmonized$cancer_type
)

# Flag which cancer types were inferred
metadata_harmonized$paipu_cancer_type_inferred <- 
  is_generic_cancer(metadata_harmonized$cancer_type) &
  metadata_harmonized$paipu_tissue_final != "unlabeled"

```

```{r}
# See how many cancer types were inferred from tissue
sum(is_generic_cancer(metadata_harmonized$cancer_type) &
  metadata_harmonized$paipu_tissue_final != "unlabeled"
)

```

## Drop:
### Macaca dascicularis BioProject PRJNA603354 - diet
### Bos taurus PRJNA273618 - Cooperia oncophora
### Macaca mulatta PRJNA694855 - infections
### Rattus norvegicus BioProject PRJNA808966 - diabetes 
### Rattus norvegicus PRJNA507730 - immortalized Muller cell lines
### Rattus norvegicus PRJNA540257 - diets and endocrine disruptors 
### Rattus norvegicus PRJNA641512 - early life stress
### Rattus norvegicus PRJNA1111852 - stress
### Rattus norvegicus PRJNA203278 - sample info says rat, everything else says fungi
### Canis lupus familiaris PRJNA690933 - epithelial-to-mesenchymal transition (EMT)
### Canis lupus familiaris PRJNA704985 - genome architecture study (fibroblast)
### Canis lupus familiaris PRJNA338759 - MDCK kidney cell lines (not a cancer)
### Sus scrofa PRJNA629513 - alcohol liver disease
### Heterocephalus glaber PRJEB13074 - gene expression in normal liver
### SRR32133263 - normal cell strain

```{r}
metadata_harmonized <- metadata_harmonized %>%
  filter(!bioproject %in% c("PRJNA808966", "PRJNA603354", "PRJNA273618", "PRJNA694855", "PRJNA507730", "PRJNA540257", "PRJNA690933", "PRJNA704985", "PRJNA338759", "PRJNA629513", "PRJNA641512", "PRJEB13074", "PRJNA1111852", "PRJNA203278"))

metadata_harmonized <- metadata_harmonized %>%
  filter(!run_accession %in% c("SRR32133263", "SRR19329937"))

# Only the samples labeled as tumor are tumors, drop the other samples
metadata_harmonized <- metadata_harmonized %>%
  filter(!(bioproject == "PRJNA966987" & source_name != "tumor"))

# Drop the normal samples for this bioproject
metadata_harmonized <- metadata_harmonized %>%
  filter(!(bioproject == "PRJNA495060" & source_name == "Colonic normal epithelium"))


# Drop the fibroblast non-tumor cell line samples for this bioproject
metadata_harmonized <- metadata_harmonized %>%
  filter(!(bioproject == "PRJEB8646" & paipu_cancer_type == "onco*"))

metadata_harmonized <- metadata_harmonized %>%
  filter(!(bioproject == "PRJEB8646" & paipu_cancer_type == "cancer*"))
```

## Update paipu_cancer_type for certain bioprojects

```{r}
# Make all samples in PRJNA488826 -> cancer* because it doesn't specify cancer types
metadata_harmonized <- metadata_harmonized %>%
  mutate(paipu_cancer_type = ifelse(bioproject == "PRJNA488826",
                             "cancer*",
                             paipu_cancer_type))

# Make all samples in PRJNA993836, PRJNA736516 -> tumor*
metadata_harmonized <- metadata_harmonized %>%
  mutate(paipu_cancer_type = ifelse(bioproject %in% c("PRJNA993836", "PRJNA736516"),
                             "tumor*",
                             paipu_cancer_type))

# Make all samples in PRJEB8735 -> sarcoma
metadata_harmonized <- metadata_harmonized %>%
  mutate(paipu_cancer_type = ifelse(bioproject == "PRJEB8735",
                             "sarcoma",
                             paipu_cancer_type))

# Make all samples in PRJNA492338 -> mast cell tumor
metadata_harmonized <- metadata_harmonized %>%
  mutate(paipu_cancer_type = ifelse(bioproject == "PRJNA492338",
                             "mast cell tumor",
                             paipu_cancer_type))

# Make all samples in PRJDB9605 -> lung cancer
metadata_harmonized <- metadata_harmonized %>%
  mutate(paipu_cancer_type = ifelse(bioproject == "PRJDB9605",
                             "lung cancer",
                             paipu_cancer_type))

# Make all samples in PRJNA933211, PRJEB30086, PRJNA565189, PRJNA930768 -> bone cancer
metadata_harmonized <- metadata_harmonized %>%
  mutate(paipu_cancer_type = ifelse(bioproject %in% c("PRJNA933211", "PRJEB30086", "PRJNA565189", "PRJNA930768"),
                             "bone cancer",
                             paipu_cancer_type))


# Make all samples in PRJNA345482 -> osteosarcoma
metadata_harmonized <- metadata_harmonized %>%
  mutate(paipu_cancer_type = ifelse(bioproject == "PRJNA345482",
                             "osteosarcoma",
                             paipu_cancer_type))


# Make all samples in PRJNA353983 -> horn cancer
metadata_harmonized <- metadata_harmonized %>%
  mutate(paipu_cancer_type = ifelse(bioproject %in% c("PRJNA353983","PRJNA81817"),
                             "horn cancer",
                             paipu_cancer_type))

# Make all samples in PRJNA203086, PRJEB57447, PRJEB85459, PRJNA561580, PRJNA658225, PRJNA691669 -> mammary cancer
metadata_harmonized <- metadata_harmonized %>%
  mutate(paipu_cancer_type = ifelse(bioproject %in% c("PRJNA203086", "PRJEB57447", "PRJEB85459", "PRJNA561580", "PRJNA658225", "PRJNA691669", "PRJNA489087"),
                             "mammary cancer",
                             paipu_cancer_type))

# Make all 'tumor' samples in PRJNA966987 -> mammary cancer (dropped the others)
metadata_harmonized <- metadata_harmonized %>%
  mutate(paipu_cancer_type = ifelse(bioproject == "PRJNA966987" & source_name == "tumor",
                            "mammary cancer",
                            paipu_cancer_type))

# Make all samples in PRJEB19508, PRJEB21960 -> canine transmissible venereal tumor
metadata_harmonized <- metadata_harmonized %>%
  mutate(paipu_cancer_type = ifelse(bioproject %in% c("PRJEB19508", "PRJEB21960"),
                             "canine transmissible venereal tumor",
                             paipu_cancer_type))

# Make all samples in PRJNA601534 -> pheochromocytoma
metadata_harmonized <- metadata_harmonized %>%
  mutate(paipu_cancer_type = ifelse(bioproject == "PRJNA601534",
                             "pheochromocytoma",
                             paipu_cancer_type))

# Make all samples in PRJNA671877 -> oligodendroglioma
metadata_harmonized <- metadata_harmonized %>%
  mutate(paipu_cancer_type = ifelse(bioproject == "PRJNA671877",
                             "oligodendroglioma",
                             paipu_cancer_type))

# Make all samples in PRJNA376380, PRJNA562916 -> hemangiosarcoma
metadata_harmonized <- metadata_harmonized %>%
  mutate(paipu_cancer_type = ifelse(bioproject %in% c("PRJNA376380", "PRJNA562916"),
                             "hemangiosarcoma",
                             paipu_cancer_type))

# Make all samples in PRJNA803356 -> adrenocortical tumor
metadata_harmonized <- metadata_harmonized %>%
  mutate(paipu_cancer_type = ifelse(bioproject == "PRJNA803356",
                             "adrenocortical tumor",
                             paipu_cancer_type))

# Make all samples in PRJNA904320, PRJNA305035 -> hepatocellular carcinoma
metadata_harmonized <- metadata_harmonized %>%
  mutate(paipu_cancer_type = ifelse(bioproject %in% c("PRJNA904320", "PRJNA305035", "PRJNA668800"),
                             "hepatocellular carcinoma",
                             paipu_cancer_type))

# Make all samples in PRJNA475196 -> oral melanoma
metadata_harmonized <- metadata_harmonized %>%
  mutate(paipu_cancer_type = ifelse(bioproject == "PRJNA475196",
                             "oral melanoma",
                             paipu_cancer_type))

# Make all samples in PRJNA510994 -> papilloma
metadata_harmonized <- metadata_harmonized %>%
  mutate(paipu_cancer_type = ifelse(bioproject == "PRJNA510994",
                             "papilloma",
                             paipu_cancer_type))

# Make all samples in PRJNA531174 -> leiomyoma
metadata_harmonized <- metadata_harmonized %>%
  mutate(paipu_cancer_type = ifelse(bioproject == "PRJNA531174",
                             "leiomyoma",
                             paipu_cancer_type))

# Make all samples in PRJNA268531 -> lymphoma
metadata_harmonized <- metadata_harmonized %>%
  mutate(paipu_cancer_type = ifelse(bioproject == "PRJNA268531",
                             "lymphoma",
                             paipu_cancer_type))


```

## Create a final cancer type column for grouping

```{r}
metadata_harmonized <- metadata_harmonized %>%
  mutate(
    paipu_cancer_type_final = case_when(
      paipu_cancer_type %in% c("breast cancer", "mammary cancer", "mammary carcinoma", "ductal carcinoma", "pagets disease of the breast")
      ~ "breast cancer",
      
      paipu_cancer_type == "oral melanoma" ~ "melanoma",
      
      paipu_cancer_type == "horn cancer" ~ "squamous cell carcinoma",
      
      paipu_cancer_type == "large intestine cancer" ~ "colorectal cancer",
      
      paipu_cancer_type %in% c("hepatocellular cancer", "hepatocellular carcinoma") 
      ~ "liver cancer",
      
      paipu_cancer_type %in% c("soft tissue sarcoma", "hemangiosarcoma", "osteosarcoma", "rhabdomyosarcoma") 
      ~ "sarcoma",
      
      paipu_cancer_type %in% c("glioblastoma", "glioma", "oligodendroglioma", "meningioma") 
      ~ "brain cancer",
      
      # If no mapping exists, use the original name from paipu_cancer_type
      TRUE ~ paipu_cancer_type
    )
  )
```

```{r}
mammary_runs <- c(
  "SRR13200358","SRR13200359","SRR13200360","SRR13200361",
  "SRR13200362","SRR13200363","SRR13200364","SRR13200365",
  "SRR13200366","SRR13200367","SRR13200368","SRR13200369",
  "SRR13200370","SRR13200371","SRR13200372","SRR13200373",
  "SRR13200374","SRR13200375","SRR13200376","SRR13200377",
  "SRR13200378","SRR13200379"
)

prostate_runs <- c(
  "SRR17156393","SRR17156394","SRR17156395","SRR17156396",
  "SRR17156405","SRR17156406","SRR17156407","SRR17156408",
  "SRR17156409","SRR17156410","SRR17156411","SRR17156412",
  "SRR17156413","SRR17156414","SRR17156415","SRR17156416",
  "SRR17156417","SRR17156418","SRR17156419","SRR17156420",
  "SRR17156421","SRR17156422","SRR17156423","SRR17156424",
  "SRR17156425","SRR17156426","SRR17156427","SRR17156428",
  "SRR17156429","SRR17156430","SRR17156431","SRR17156432",
  "SRR17156445","SRR17156446","SRR17156447","SRR17156448",
  "SRR17156449","SRR17156450","SRR17156451","SRR17156452",
  "SRR17156453","SRR17156454","SRR17156455","SRR17156456",
  "SRR17156457","SRR17156458","SRR17156459","SRR17156460",
  "SRR17156461","SRR17156462","SRR17156463","SRR17156464",
  "SRR17156477","SRR17156478","SRR17156479","SRR17156480",
  "SRR17156481","SRR17156482","SRR17156483","SRR17156484",
  "SRR17156501","SRR17156502","SRR17156503","SRR17156504",
  "SRR17156505","SRR17156506","SRR17156507","SRR17156508"
)
```


## Update paipu_cancer_type and paipu_cancer_type_final

```{r}

metadata_harmonized <- metadata_harmonized %>%
  mutate(
    paipu_cancer_type = case_when(
      
      # Breast
      bioproject %in% c("PRJNA1014121", "PRJNA383059", "PRJNA557680") |
        run_accession %in% mammary_runs ~ "mammary carcinoma",
      
      # Colon / colorectal
      run_accession == "SRR12033544" ~ "colonic adenocarcinoma",
      bioproject == "PRJNA940677" ~ "colon carcinoma",
      
      # Endometrial
      bioproject == "PRJNA378639" ~ "endometrial adenocarcinoma",
      
      # Kidney
      bioproject == "PRJNA597461" ~ "renal cell carcinoma",
      
      # Liver
      paipu_tissue_final %in% c("liver", "liver cell line") &
        paipu_cancer_type == "carcinoma" ~ "hepatocellular carcinoma",
      
      # Oral squamous cell carcinoma
      bioproject %in% c("PRJNA533473", "PRJEB34234", "PRJNA260600") ~ "oral squamous cell carcinoma",
        
      # Pancreatic
      paipu_tissue_final == "pancreas cell line" ~ "insulinoma",
      
      # Prostate
      bioproject %in% c("PRJNA683222", "PRJNA786902") ~ "prostate adenocarcinoma",
      
      bioproject %in% c("PRJEB47028", "PRJNA506907") ~ "prostate carcinoma",
      
      # Thyroid
      bioproject == "PRJEB43017" ~ "thyroid follicular cell carcinoma",
      
      # Otherwise keep original labels
      TRUE ~ paipu_cancer_type
    )
  )

metadata_harmonized <- metadata_harmonized %>%
  mutate(
    paipu_cancer_type_final = case_when(
      # Bladder
      paipu_tissue_final %in% c("bladder", "bladder cell line") & 
        paipu_cancer_type_final != "bladder cancer" ~ "bladder cancer",
      
      # Breast
      bioproject %in% c("PRJNA1014121", "PRJNA383059", "PRJNA557680", "PRJNA912710") |
        run_accession %in% mammary_runs ~ "breast cancer",
      
      bioproject == "PRJNA558957" ~ "mammary adenoma",
      
      
      # Colon / colorectal
      run_accession == "SRR12033544" ~ "colorectal cancer",
      bioproject == "PRJNA940677" ~ "colorectal cancer",
      bioproject == "PRJNA495060" ~ "colonic adenoma",
      
      # Endometrial
      bioproject == "PRJNA378639" ~ "endometrial cancer",
      
      # Kidney
      bioproject == "PRJNA597461" ~ "kidney cancer",
      
      # Liver
      paipu_tissue_final %in% c("liver", "liver cell line") & 
        paipu_cancer_type_final != "liver cancer" &
        !bioproject %in% c("PRJNA562916", "PRJNA376380") ~ "liver cancer",
      
      # Oral squamous cell carcinoma
      bioproject %in% c("PRJNA533473", "PRJEB34234", "PRJNA260600") ~ "oral squamous cell carcinoma",
      
      # Pancreatic
      paipu_tissue_final == "pancreas" & 
        paipu_cancer_type_final != "pancreatic cancer" ~ "pancreatic cancer",
      
      paipu_tissue_final == "pancreas cell line" ~ "pancreatic neuroendocrine tumor",
      
      # Prostate
      bioproject %in% c("PRJNA683222", "PRJNA786902", "PRJEB47028", "PRJNA506907") | 
        run_accession %in% prostate_runs ~ "prostate cancer",
      
      # Thyroid
      bioproject == "PRJEB43017" ~ "thyroid cancer",
      
      # Otherwise keep original labels
      TRUE ~ paipu_cancer_type_final
    )
  )
```


```{r}
# Check unique paipu_cancer_type_final now
length(unique(metadata_harmonized$paipu_cancer_type_final))
table(metadata_harmonized$paipu_cancer_type_final)
```

# Harmonize tumor site

```{r}
grep("tumor.*site|tumor.*loc|tumor_location|tumor_site|sampling_site|tumor_origin", names(metadata_harmonized), ignore.case = TRUE, value = TRUE)

```

```{r}
# Apply first_nonempty() function to create a harmonized breed col
metadata_harmonized$paipu_tumor_site <- mapply(
  first_nonempty,
  metadata_harmonized$tumor_location,
  metadata_harmonized$sampling_site
)

# Check
unique(metadata_harmonized$paipu_tumor_site)
```

# Harmonize treatment 
```{r}
# Check cols that have drug, treatment, etc
grep("treat|drug|compound|therapy|intervention", names(metadata_harmonized), ignore.case = TRUE, value= TRUE)

treatment_cols <- c(
  "treatment",
  "treatment_group",
  "compound",
  "intervention"
)
```

```{r}
# Apply first_nonempty() function to create a harmonized treatment col
metadata_harmonized$paipu_treatment <- do.call(
  function(...) mapply(first_nonempty, ..., USE.NAMES = FALSE),
  metadata_harmonized[treatment_cols]
)

# Create a flag where more than one tumor site col has data
treatment_candidates <- metadata_harmonized[treatment_cols]

# Count how many tissue cols are filled per row
treatment_filled_count <- rowSums(
  !is.na(treatment_candidates) & trimws(as.matrix(treatment_candidates)) != ""
)

metadata_harmonized$paipu_treatment_multiple_sources <- treatment_filled_count > 1 # true if more than one col had a cell type

# Check how many rows have more than 1 source (0)
table(metadata_harmonized$paipu_treatment_multiple_sources) 

```

# Harmonize timepoint

```{r}
# Check cols that have time point
grep("time|point", names(metadata_harmonized), ignore.case = TRUE, value = TRUE)

```

```{r}
# Check if any conflicts (if > 0, then conflict)
sum(!is.na(metadata_harmonized$timepoint) & 
    !is.na(metadata_harmonized$time_point) & 
    metadata_harmonized$timepoint != metadata_harmonized$time_point)
```

```{r}
# Combine the timepoint cols
metadata_harmonized <- metadata_harmonized %>%
  mutate(paipu_time_point = coalesce(timepoint, time_point))

unique(metadata_harmonized$paipu_time_point)
```

# Harmonize time of sampling

```{r}
# Check if any conflicts
sum(!is.na(metadata_harmonized$time_of_sample) & 
    !is.na(metadata_harmonized$time_of_sampling) & 
    metadata_harmonized$time_of_sample != metadata_harmonized$time_of_sampling)
```

```{r}
# Combine the timepoint cols
metadata_harmonized <- metadata_harmonized %>%
  mutate(paipu_time_of_sampling = coalesce(time_of_sample, time_of_sampling))

unique(metadata_harmonized$paipu_time_of_sampling)
```

# Harmonize age

```{r}
# Check cols that have age
grep("age", names(metadata_harmonized), ignore.case = TRUE, value = TRUE)

```

```{r}
# Combine dev stage cols
metadata_harmonized <- metadata_harmonized %>%
  mutate(
    paipu_age = coalesce(age, age_of_dog, estimated_age)
  )
```


# Harmonize development stage

```{r}
# Combine dev stage cols
metadata_harmonized <- metadata_harmonized %>%
  mutate(
    paipu_development_stage = coalesce(development_stage, developmental_stage, dev_stage)
  )

table(metadata_harmonized$paipu_development_stage)
```

```{r}
# Fix misspelling of adult
metadata_harmonized$paipu_development_stage[
  metadata_harmonized$paipu_development_stage %in% c("adlt", "adult")] <- "Adult"

unique(metadata_harmonized$paipu_development_stage)
```

# Group cancer types by location/system HERE TODO START HERE
```{r}
unique(metadata_harmonized$paipu_cancer_type_final)
```
```{r}
# Set cancer type systems for each cancer_type
blood <- c("lymphoma", "leukemia") # Hematologic/Blood 

breast <- c("breast cancer", "mammary adenoma")

digestive <- c("liver cancer","pancreatic cancer", "colorectal cancer", "pancreatic neuroendocrine tumor", "colonic adenoma")

endocrine <- c("pheochromocytoma", "pituitary tumor", "adrenocortical tumor") # Endocrine and Neuroendocrine

eye <- c("eye cancer")

genitourinary <- c("bladder cancer", "prostate cancer", "canine transmissible venereal tumor", "kidney cancer")

gynecologic <- c("endometrial cancer",  "choriocarcinoma")

head_neck <- c("oral squamous cell carcinoma", "thyroid cancer")

musculoskeletal <- c("sarcoma", "bone cancer")

neurologic <- c("brain cancer", "neuroblastoma")
       
respiratory <- c("lung cancer", "mesothelioma")     

skin <- c("melanoma", "histiocytoma", "mast cell tumor")

unspecified <- c("tumor*", "cancer*")

# Create a cancer_system column
metadata_harmonized <- metadata_harmonized %>%
  mutate(paipu_cancer_system = case_when(
    paipu_cancer_type_final %in% blood ~ "Hematologic_Blood",
    paipu_cancer_type_final %in% breast ~ "Breast",
    paipu_cancer_type_final %in% digestive ~ "Digestive_Gastrointestinal",
    paipu_cancer_type_final %in% endocrine ~ "Endocrine_Neuroendocrine",
    paipu_cancer_type_final %in% eye ~ "Eye",
    paipu_cancer_type_final %in% genitourinary ~ "Genitourinary",
    paipu_cancer_type_final %in% gynecologic ~ "Gynecologic",
    paipu_cancer_type_final %in% head_neck ~ "Head and Neck",
    paipu_cancer_type_final %in% musculoskeletal ~ "Musculoskeletal",
    paipu_cancer_type_final %in% neurologic ~ "Neurologic",
    paipu_cancer_type_final %in% respiratory ~ "Respiratory_Thoracic",
    paipu_cancer_type_final %in% skin ~ "Skin",
    paipu_cancer_type_final %in% unspecified ~ "Unspecified"
  ))
                                

# Check the new col
table(metadata_harmonized$paipu_cancer_system)
```

```{r}
# metadata_harmonized %>%
#   filter(paipu_cancer_type_final == "tumor*")
```


```{r}
# Create a manual curation note column to store notes on manual labels for ductal carcinoma
metadata_harmonized$manual_curation_note <- NA_character_


# Label cancer system for ductal carcinoma based on other metadata info
metadata_harmonized <- metadata_harmonized %>%
  mutate(
    paipu_cancer_system = case_when(

      # manual - ductal carcinoma in these bioprojs is pancreatic (from NCBI SRA)
      paipu_cancer_type == "ductal carcinoma" &
        bioproject %in% c("PRJNA1125449", "PRJNA1124987") ~
        "Digestive_Gastrointestinal",
      
      # leiomyoma in PRJNA531174 involves lymphangioleiomyomatosis (LAM), so respiratory
      bioproject == "PRJNA531174" ~ "Respiratory_Thoracic",
      
      # source in PRJNA488826 is peripheral blood; tissue is peripheral blood mononuclear cells
      bioproject == "PRJNA488826" ~ "Hematologic_Blood",
      
      # papilloma in PRJNA510994 is wart tissue and route/site is cutaneous
      bioproject == "PRJNA510994" ~ "Skin",
      
      # adenoma in PRJNA1089080 is hepatoid perianal glands
      # squamous cell carcinoma in PRJNA81817, PRJNA353983, PRJNA497667 is horn cancer
      bioproject %in% c("PRJNA1089080", "PRJNA81817", "PRJNA353983","PRJNA497667" ) ~ "Skin",
      
      # PRJNA736516 has samples with tumors in different parts, but no specific cancer type
        # SRR14773911, SRR14773912, SRR14773913 - nose
        # SRR14773914, SRR14773915, SRR14773916 - mouth
        # SRR14773917, SRR14773918, SRR14773919 - scrotum
      run_accession %in% c("SRR14773911", "SRR14773912", "SRR14773913") ~ "Head and Neck",
      run_accession %in% c("SRR14773914", "SRR14773915", "SRR14773916") ~ "Head and Neck",
      run_accession %in% c("SRR14773917", "SRR14773918", "SRR14773919") ~ "Genitourinary",
      
      # ductal carcinoma + breast/mammary in other metadata cols
      paipu_cancer_type == "ductal carcinoma" &
        str_detect(
          paste(disease, paipu_cell_type, sep = " "),
          "breast|mammary"
        ) ~ "Breast",

      # ductal carcinoma + pancreatic/liver in other metadata cols
      paipu_cancer_type == "ductal carcinoma" &
        str_detect(
          paste(disease, paipu_cell_type, sep = " "),
          "pancreas|pancreatic|liver|hepatic"
        ) ~ "Digestive_Gastrointestinal",

      TRUE ~ paipu_cancer_system # if no matches, keep what cancer_system had
    ),
    
    manual_curation_note = case_when(
      paipu_cancer_type == "ductal carcinoma" &
        bioproject %in% c("PRJNA1125449", "PRJNA1124987") ~
        "Cancer system was assigned as Digestive_Gastrointestinal based on NCBI SRA stating Pancreatic Ductal Carcinoma in Study Title",
      
      bioproject == "PRJNA1089080" ~ "Cancer system was assigned as Skin based on NCBI SRA stating hepatoid gland adenoma (perianal)",
      
      bioproject %in% c("PRJNA81817", "PRJNA353983","PRJNA497667" ) ~ "Cancer system was assigned as Skin based on these being horn cancer",
      
      bioproject == "PRJNA510994" ~ "Cancer system was assigned as Skin based on it being wart tissue with route/site as cutaneous",
      
      bioproject == "PRJNA531174" ~ "Cancer system was assigned as Respiratory_Thoracic based on it involving lymphangioleiomyomatosis",
      
      # source in PRJNA488826 is peripheral blood; tissue is peripheral blood mononuclear cells
      bioproject == "PRJNA488826" ~ "Cancer system was assigned as Hematologic_Blood based on source being peripheral blood",
      
      # PRJNA736516 has samples with tumors in different parts, but no specific cancer type
        # SRR14773911, SRR14773912, SRR14773913 - nose
        # SRR14773914, SRR14773915, SRR14773916 - mouth
        # SRR14773917, SRR14773918, SRR14773919 - scrotum
      run_accession %in% c("SRR14773911", "SRR14773912", "SRR14773913") ~ "Cancer system was assigned as Head and Neck because tissue is nose",
      run_accession %in% c("SRR14773914", "SRR14773915", "SRR14773916") ~ "Cancer system was assigned as Head and Neck because tissue is mouth",
      run_accession %in% c("SRR14773917", "SRR14773918", "SRR14773919") ~ "Cancer system was assigned as Genitourinary because tissue is scrotum",
      
      TRUE ~ manual_curation_note
    )
  )

# Check
table(metadata_harmonized$paipu_cancer_system)
```

# Add taxonomic order to metadata

```{r}
unique(metadata_harmonized$organism_scientific_name)
```

## Rename Gorilla to Gorilla gorilla because removing repeated phrases made it 'Gorilla'
```{r}
metadata_harmonized <- metadata_harmonized %>%
  mutate(organism_scientific_name = ifelse(
    organism_scientific_name == "Gorilla",
    "Gorilla gorilla",
    organism_scientific_name
  ))
```

```{r}
# Map species to their taxonomic order
species_taxonomic_order <- c(
  # Carnivora
  "Canis lupus familiaris" = "Carnivora",
  "Enhydra lutris" = "Carnivora",
  "Felis catus" = "Carnivora",
  
  # Cetartiodactyla
  "Bos indicus" = "Cetartiodactyla",
  "Bos taurus" = "Cetartiodactyla",
  "Bubalus bubalis" = "Cetartiodactyla",
  "Sus scrofa" = "Cetartiodactyla",
  
  # Lagomorpha
  "Oryctolagus cuniculus" = "Lagomorpha",
  
  # Perissodactyla
  "Equus caballus" = "Perissodactyla",
  
  # Primates
  "Chlorocebus sabaeus" = "Primates",
  "Gorilla gorilla" = "Primates",
  "Macaca fascicularis" = "Primates",
  "Macaca mulatta" = "Primates",
  "Pan troglodytes" = "Primates",
  
  # Rodentia
  "Cricetulus griseus" = "Rodentia",
  "Mesocricetus auratus" = "Rodentia",
  "Rattus norvegicus" = "Rodentia",
  "Heterocephalus glaber" = "Rodentia",
  
  # Scandentia
  "Tupaia chinensis" = "Scandentia"
)
```

```{r}
# Create a clade col
metadata_harmonized$clade <- species_taxonomic_order[metadata_harmonized$organism_scientific_name]

# Check
table(metadata_harmonized$clade)
```

# Drop all columns that are completely empty

```{r}
# Find columns where every value is NA
na_cols <- sapply(metadata_harmonized, function(x) all(is.na(x)))

metadata_harmonized[, na_cols] # print NA cols

# Count number of NA cols
sum(na_cols)

# Remove NA cols
metadata_harmonized <- metadata_harmonized[, !na_cols]

```

# Drop harmonization cols

```{r}
# Cols to drop for harmonized metadata
cols_to_drop <- c("gender", "gender_animal_site", "reproductive_status", "sex", "neuter_status", "animal", "spay_neuter_status", "neutering", "reproductive_status", "tumor_grade", "tumor_grading", "infection_1", "infection_2", "infection", "infect", "fcov_infection", "cell_type", "cell_type.1", "cell_type.3", "injected_cell_type", "tissue_cell_type_source", "breed", "dog_breed", "tissue", "organ", "organism_part", "tissuetype", "tissue_type", "tissue_source", "treatment", "treatment_group", "intervention", "sampling_site", "tumor_location", "timepoint", "time_point", "time_of_sampling", "time_of_sample", "age", "age_of_dog",  "development_stage", "developmental_stage", "dev_stage")
#"cell_type.2","compound", "estimated_age",

metadata_harmonized <- metadata_harmonized %>%
  select(-all_of(cols_to_drop))

```

```{r}
ncol(metadata_harmonized)
```

# Output harmonized species metadata (duplicated)

```{r}
# Save as RDS
saveRDS(metadata_harmonized, file = here("output_data/non_mouse_data/updated_data/species_full_metadata_harmonized_5-2-26.rds"))

write_tsv(metadata_harmonized, file = here("output_data/non_mouse_data/updated_data/species_full_metadata_harmonized_5-2-26.tsv"))
```

# Drop Paipu unnecessary cols to produce final metadata 

```{r}
# Check cols with 'paipu' in the name
grep("paipu", names(metadata_harmonized), value = TRUE)
```

```{r}
# Cols to drop for final metadata
paipu_cols_to_drop <- c("cancer_type", "paipu_tissue", "paipu_cell_type_multiple_sources", "paipu_tissue_multiple_sources", "paipu_tissue_explicit", "paipu_tissue_inferred", "paipu_cancer_type_inferred", "paipu_treatment_multiple_sources", "manual_curation_note")

metadata_harmonized_final <- metadata_harmonized %>%
  select(-all_of(paipu_cols_to_drop))

```

```{r}
ncol(metadata_harmonized_final)
```

## Output final harmonized species metadata (duplicated)

```{r}
# Save as RDS
saveRDS(metadata_harmonized_final, file = here("output_data/non_mouse_data/updated_data/final_metadata/species_final_dup_metadata_5-2-26.rds"))

write_tsv(metadata_harmonized_final, file = here("output_data/non_mouse_data/updated_data/final_metadata/species_final_dup_metadata_5-2-26.tsv"))
```



