---
title: "Metadata Charts and Statistics"
output: html_document
---

```{r}
library(dplyr)
library(ggplot2)
library(forcats)
library(patchwork)
library(here)
```

# Charts

```{r}
## Species colors by taxonomic group
# Reference:
# base_colors <- c(
#   "Carnivora" = "#C02020",      # red
#   "Cetartiodactyla" = "#B878F8", # purple
#   "Rodentia" = "#2E3C90",       # dark blue
#   "Primates" = "#009E73",       # green
#   "Perissodactyla" = "#FFA500", # orange
#   "Lagomorpha" = "#DA1884",     # pink
#   "Scandentia" = "#4DBBD5"      # light blue
# )

# New species colors based on taxonomic group
species_colors <- c(
  # Carnivora
  "Canis lupus familiaris" = "#C02020",
  "Enhydra lutris" = "#CD4D4D",
  "Felis catus" = "#861616",
  
  # Cetartiodactyla
  "Bos indicus" = "#6C0BCE",
  "Bos taurus" = "#6E4895",
  "Bubalus bubalis" = "#D4AEFB",
  "Sus scrofa" = "#B878F8",
  
  # Lagomorpha
  "Oryctolagus cuniculus" = "#DA1884",
  
  # Perissodactyla
  "Equus caballus" = "#FFA500",
  
  # Primates
  "Chlorocebus sabaeus" = "#00664D",
  #"Gorilla gorilla" = "#003325",
  #"Macaca fascicularis" = "#00D49B",
  "Macaca mulatta" = "#008000",
  "Pan troglodytes" = "#66C5AB",
  
  # Rodentia
  "Cricetulus griseus" = "#6E7CD1",
  "Mesocricetus auratus" = "#2E3C90",
  "Heterocephalus glaber" = "#4C5BB5",
  "Rattus norvegicus" = "#9EA8E0",
  
  # Scandentia
  "Tupaia chinensis" = "#4DBBD5"
)

```


```{r}
# Read in the harmonized metadata
harmonized_metadata <- readRDS(here("output_data/non_mouse_data/updated_data/species_no_dup_metadata_harmonized_5-3-26.rds"))

```

## Create a legend pdf with all species excluding mouse
```{r}
# Species for the legend
legend_species <- sort(names(species_colors))

# Dummy df just for legend
legend_df <- data.frame(
  organism_scientific_name = factor(legend_species, levels = legend_species)
)

# Plot only with legend
legend_plot <- ggplot(
  legend_df,
  aes(x = 1, y = organism_scientific_name, fill = organism_scientific_name)
) +
  geom_tile() +
  scale_fill_manual(values = species_colors, drop = FALSE) +
  guides(fill = guide_legend(title = "Species", # legend title
                                ncol=1)) + # 1 column legend
  theme_void() + # remove all plot elements
  theme(
    legend.position = "right",
    legend.text = element_text(size = 18),
    legend.title = element_text(size = 19)
  )

legend_plot
```

```{r}
library(cowplot) # to save legend by itself

legend_only <- get_legend(legend_plot) # get the legend

ggsave(
  filename = here("plots/updated_data/species_legend_no_mouse_one_col.pdf"),
  plot = legend_only,
  width = 4,
  height = 5
)
```

## Sample, cancer type and species count

```{r}
nrow(harmonized_metadata)
length(unique(harmonized_metadata$organism_scientific_name))
length(unique(harmonized_metadata$paipu_cancer_type_final))
```


## Count unique bioprojects and samples

```{r}
library(data.table)

uniqueN(harmonized_metadata, by = "bioproject")
uniqueN(harmonized_metadata, by = "run_accession")
```

```{r}
# Average number of samples per bioproject
bioproj_summary <- harmonized_metadata %>%
  group_by(bioproject) %>%
  summarise(n_samples = n())

print(paste0("Mean: ", mean(bioproj_summary$n_samples)))
print(paste0("Median: ", median(bioproj_summary$n_samples)))
print(paste0("SD: ", sd(bioproj_summary$n_samples)))
print(paste0("IQR ", quantile(bioproj_summary$n_samples, probs = c(0.25, 0.75))))

```

```{r}
print(paste0("The combined data atlas is ", round(3484 / 17.4,0)," times larger than the average BioProject"))
```

## Calculate total run_size and min, max, median, mean of run_size

```{r}
# Check if any NAs
any(is.na(harmonized_metadata$run_size))
```

```{r}
# Functions to convert bytes to MB, GB and TB
bytes_to_kb <- function(x) x / 1024
bytes_to_mb <- function(x) x / 1024^2
bytes_to_gb <- function(x) x / 1024^3
bytes_to_tb <- function(x) x / 1024^4

# Get all run sizes (in bytes) excluding NAs
run_size_bytes <- harmonized_metadata$run_size
run_size_bytes <- run_size_bytes[!is.na(run_size_bytes)]

#str(run_size_bytes) # check type

# Convert run size to numeric
run_size_bytes <- as.numeric(run_size_bytes)
```

```{r}
# Run size summary
print(paste0("Total data processed: ", round(bytes_to_tb(sum(run_size_bytes)), 2), "TB"))

print(paste0("Number of runs with size: ", length(run_size_bytes)))
print(paste0("Number of runs total: ", nrow(harmonized_metadata)))


print(paste0("Min run size: ", round(bytes_to_gb(min(run_size_bytes)), 2), "GB"))
print(paste0("Min run size: ", round(bytes_to_kb(min(run_size_bytes)), 2), " KB"))

print(paste0("Max run size: ", round(bytes_to_gb(max(run_size_bytes)), 2), "GB"))

print(paste0("Median run size: ", round(bytes_to_gb(median(run_size_bytes)), 2), "GB"))

print(paste0("Mean run size: ", round(bytes_to_gb(mean(run_size_bytes)), 2), "GB"))

```

## Bar plots for given categories

```{r}
# Set threshold to combine categories into 1 if there are more than this many unique levels
default_max_levels <- 30

# Function to combine unique labels, if needed
conditionally_combine <- function(x, max_levels) { 
  # If max_levels is Inf, keep all levels as is
  if (is.infinite(max_levels)) return(as.factor(x))
  
  # Count number of unique values
  n_levels <- n_distinct(x, na.rm = TRUE) # don't consider NAs
  
  if (n_levels > max_levels) {
    # Keep top 20 levels, combine remaining levels into other
    fct_lump_n(as.factor(x), n = max_levels) 
  } else {
    as.factor(x) # otherwise, keep all levels as is
  }
}

```

```{r}
# Function to make bar plots for each given category (all species or no mouse)
make_species_barplot <- function(df, x_col, title, x_label = NULL, max_levels = 25, flip_axes = FALSE, all_caps = FALSE, y_breaks = NULL) {

  # Only keep non-NA rows
  plot_data <- df %>% filter(!is.na(.data[[x_col]])) 
  
  # Combine labels if the threshold is exceeded
  plot_data[[x_col]] <- conditionally_combine(plot_data[[x_col]], max_levels)

  # Order labels by frequency
  plot_data[[x_col]] <- fct_infreq(plot_data[[x_col]])
  
  # Let "Other" be at the end if labels were combined
  if ("Other" %in% levels(plot_data[[x_col]])) { 
    plot_data[[x_col]] <- fct_relevel(plot_data[[x_col]], "Other", after = Inf) 
  }
  
  # Put most frequent label at the top when axes are flipped
  if (flip_axes) { 
    plot_data[[x_col]] <- fct_rev(plot_data[[x_col]])
  }

  # Replace underscores with spaces for plots
  levels(plot_data[[x_col]]) <- gsub("_", " ", levels(plot_data[[x_col]]))
  
  # If flag is true, make all labels be all caps and keep their order
  if (all_caps) {
    plot_data[[x_col]] <- factor(
      toupper(as.character(plot_data[[x_col]])),
      levels = toupper(levels(plot_data[[x_col]])))
  }
    
  # Create plot
  p <- ggplot(plot_data, aes(x = .data[[x_col]], fill = organism_scientific_name)) +
    geom_bar() + # draw count bars
    scale_fill_manual(values = species_colors, drop = FALSE) + # use the species color palette
    theme_bw() + # clean black and white theme
    labs(
      x = ifelse(is.null(x_label), x_col, x_label), # use x label (if given in function call) or default to col name
      y = "Number of samples",
      fill = "Species",
      title = title
    ) + 
    theme( # Increase font size
      axis.title.x = element_blank(), # remove x-axis title
      axis.text = element_text(size = 16), # font size for all axes
      axis.title.y = element_text(size = 16), # font size for "Number of samples"
      plot.title = element_text(size = 18, hjust = 0.5, face = "bold") # make titles bold and centered
  )
  
  # Set y-axis marks (optional)
  if (!is.null(y_breaks)) {
    max_count <- max(table(plot_data[[x_col]]))
    
    p <- p + scale_y_continuous(
      breaks = y_breaks,
      limits = c(0, max(max(y_breaks), max_count))
    )
  }
  
  if (flip_axes) {
    p <- p + coord_flip() # flip axes so labels are readable
  } else {
    p <- p + theme(axis.text.x = element_text(angle = 45, hjust = 1)) # rotate x-axis labels
  }
  p # return plot
}

# Function to save bar plots (png and pdf)
save_plot_png_pdf <- function(p, output_name, width = 9, height = 6, dpi = 300) {
  # Define the fig path
  fig_path <- here(paste0("plots/updated_data/", output_name))
  
  ggsave(filename = paste0(fig_path, ".pdf"), plot = p, width = width, height = height, device = cairo_pdf)
}

```


## Create charts to make 2 figures

```{r}
# Non-mouse technical metadata plots (figure 1)
library_layout_plot <- make_species_barplot(harmonized_metadata, "library_layout", "Library Layout", "Library Layout", default_max_levels,all_caps = TRUE, y_breaks = c(0, 1000, 2000, 3000))

single_bulk_plot <- make_species_barplot(harmonized_metadata, "single_bulk", "Bulk vs Single-Cell", "Bulk vs Single-Cell", default_max_levels, all_caps = TRUE, y_breaks = c(0, 1000, 2000, 3000))

sex_plot <- make_species_barplot(harmonized_metadata, "paipu_sex_final", "Sex", "Sex", default_max_levels, all_caps = TRUE, y_breaks = c(0, 1000, 2000, 3000)) 

platform_plot <- make_species_barplot(harmonized_metadata, "platform_instrument_model", "Sequencing Platform", "Sequencing Platform", default_max_levels, all_caps = TRUE) 

```

```{r}
# Biological metadata plots (figure 2)
# Top / bottom charts
cancer_type_plot_top <- make_species_barplot(harmonized_metadata, "paipu_cancer_type_final", "Cancer Type", "Cancer Type", max_levels = Inf)

tissue_plot_top <- make_species_barplot(harmonized_metadata, "paipu_tissue_final", "Tissue", "Tissue", max_levels = Inf)
```

```{r}
# Remove all legends from plots
library_layout_plot <- library_layout_plot + theme(legend.position = "none", axis.text = element_text(color = "black"))

single_bulk_plot <- single_bulk_plot + theme(legend.position = "none", axis.text = element_text(color = "black"))

sex_plot <- sex_plot + theme(legend.position = "none", axis.text = element_text(color = "black"))

platform_plot <- platform_plot + theme(legend.position = "none", plot.margin = margin(l = 35), axis.text = element_text(color = "black")) # extra space on the left to avoid cutting off x-axis labels

# Top / bottom chart
cancer_type_plot_top <- cancer_type_plot_top + theme(legend.position = "none", plot.title = element_text(size = 26, hjust = 0.5, face = "bold"), axis.text = element_text(size = 22, color = "black"), axis.title.x = element_blank(), axis.title.y = element_text(size = 22, angle = 90, margin = margin(r=15)))

tissue_plot_top <- tissue_plot_top + theme(legend.position = "none", plot.title = element_text(size = 26, hjust = 0.5, face = "bold"), axis.text = element_text(size = 22, color = "black"), axis.title.x = element_blank(), axis.title.y = element_text(size = 22, angle = 90, margin = margin(r=15))) # margin adds space between y-axis label and y-axis numbers
      
```


```{r}
# Create figure 1: bulk vs single, library layout and sex across the top and sequencing platform across the bottom with space on the right of it for legend
top_row <- single_bulk_plot | library_layout_plot | sex_plot
bottom_row <- (platform_plot | plot_spacer()) + plot_layout(widths = c(3, 1)) # make platform cover part of the area to leave room to add legend

# Put rows together for the figure
fig1 <- top_row / bottom_row + plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 22)) # increase panel font sizes

# Create figure 2 [paipu_cancer_type_final]
fig2_ <- (cancer_type_plot_top / tissue_plot_top) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 22)) # increase panel font sizes

```

```{r}
# Save figures as pdfs
ggsave(here("plots/updated_data/fig1_tech_metadata.pdf"), fig1, width = 16, height = 15, device = cairo_pdf)

ggsave(here("plots/updated_data/fig2_biological_metadata_harmonized_top.pdf"), fig2, width = 27, height = 20, device = cairo_pdf)

```

# Statistics

## Species data percentage

```{r}
prop.table(table(harmonized_metadata$organism_scientific_name) *100)

```

## Which clade has the most species represented?

```{r}
# Find which taxonomic order has the most species represented
harmonized_metadata %>%
  distinct(organism_scientific_name, clade) %>%
  count(clade, sort = TRUE)
```


## Calculate median, minimum and maximum of run_total_spots for sequencing depth
```{r}
# Check if any NAs
table(is.na(harmonized_metadata$run_total_spots))

# Check if the col is numeric
is.numeric((harmonized_metadata$run_total_spots))

# Make run_total_spots numeric
harmonized_metadata$run_total_spots <- as.numeric(harmonized_metadata$run_total_spots)

# Check
is.numeric((harmonized_metadata$run_total_spots))

```

```{r}
# Calculate median, min and max in millions
median_spots <- median(harmonized_metadata$run_total_spots)/1000000
min_spots <- min(harmonized_metadata$run_total_spots)/1000000
max_spots <- max(harmonized_metadata$run_total_spots)/1000000

print(paste0("Samples had a median sequencing depth of ", round(median_spots, 1), " million spots per sample with a range of ", round(min_spots, 1), "-", round(max_spots, 1), " million."))

# True minimum
print(min_spots*1000000)
```


## Sequencing reads by species
```{r}
# Compute total reads (paired *2, single as is)
harmonized_metadata$total_reads_test <- ifelse(harmonized_metadata$library_layout == "PAIRED", harmonized_metadata$run_total_spots * 2, harmonized_metadata$run_total_spots)

```

```{r}
# Kruskal-Wallis test
kruskal.test(harmonized_metadata$total_reads_test, harmonized_metadata$organism_scientific_name)
```

## Species vs library layout
```{r}
species_layout_table <- table(harmonized_metadata$organism_scientific_name, harmonized_metadata$library_layout)

species_layout_table
```

```{r}
# Check proportions
prop.table(species_layout_table,1)
```

```{r}
# Check expected values for if they're < 5
chisq.test(species_layout_table)$expected
```

```{r}
# Use simulated p-value because there are small counts
chisq.test(species_layout_table, simulate.p.value = TRUE, B = 10000)
```

```{r}
# Check if the significance still exists without dog and rat
non_mouse_dog_rat_metadata <- subset(harmonized_metadata,
                        !organism_scientific_name %in% c("Canis lupus familiaris",
                                        "Rattus norvegicus"))

species_no_dog_rat_layout_table <- table(non_mouse_dog_rat_metadata$organism_scientific_name,
                    non_mouse_dog_rat_metadata$library_layout)

```

```{r}
chisq.test(species_no_dog_rat_layout_table, simulate.p.value = TRUE, B = 10000)
```



## Cancer type by species
```{r}
options(max.print = 100000)
```


```{r}
# Contingency table counting samples for each cancer type x species combo
species_ct_table <- table(harmonized_metadata$organism_scientific_name, harmonized_metadata$paipu_cancer_type_final)

```

```{r}
dim(species_ct_table)
```


```{r}
# Null hypothesis: cancer type is independent of species
# H0: P(cancer type | species) = P(cancer type)

# Alternative hypothesis: the distribution of cancer types differ across species
# H1: P(cancer type | species) != P(cancer type)

# Have to use chi-squared test with Monte Carlo simulation instead of Fisher's bc the contingency table (16 x 35) is big
chisq_results <- chisq.test(
  species_ct_table,
  simulate.p.value = TRUE, # this says compute p-vals by Monte Carlo simulation
  B = 10000 # specify the num of simulated replicates used in the Monte Carlo test
)

chisq_results
## p-val = 9.999e-05 -> reject null hypothesis (not independent of each other)
```

## Single-cell/bulk (sequencing type) vs library layout

```{r}
chisq.test(table(harmonized_metadata$single_bulk, harmonized_metadata$library_layout))
```


## Cancer malignancy prevalence

```{r}
cancer_incidence <- read.csv(here("data/zoonomia_cancer_incidence.csv"))

head(cancer_incidence)
```


```{r}
# Drop IR_Match col
cancer_incidence <- cancer_incidence %>% select(-IR_Match)

# Replace underscore in species names with a space to match metadata
cancer_incidence$Zoonomia_Species <- gsub("_", " ", cancer_incidence$Zoonomia_Species)

head(cancer_incidence)
```

```{r}
# Get num of samples per species
species_samples <- harmonized_metadata %>%
                        group_by(organism_scientific_name) %>%
                        summarise(Sample_Counts = n())

species_samples
```

```{r}
# Merge species samples and cancer incidence
sp_cancer_incidence <- inner_join(cancer_incidence, species_samples, by = c("Zoonomia_Species" = "organism_scientific_name"))

# Sort by species
sp_cancer_incidence <- sp_cancer_incidence %>% arrange(sp_cancer_incidence, Zoonomia_Species)

sp_cancer_incidence
```

```{r}
# Add a log sample count col so data isn't skewed
sp_cancer_incidence <- sp_cancer_incidence %>%
                        mutate(Log_Sample_Counts = log(Sample_Counts + 1))

sp_cancer_incidence
```

```{r}
# Spearman correlation
cor.test(sp_cancer_incidence$Malignancy_Prevalence, sp_cancer_incidence$Sample_Counts, method = "spearman")

```

```{r}
# Linear regression
model <- lm(Malignancy_Prevalence ~ Log_Sample_Counts, data = sp_cancer_incidence)
summary(model)
```

```{r}
# Plot
ggplot(sp_cancer_incidence,
       aes(x = Log_Sample_Counts,
           y = Malignancy_Prevalence)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Log Sample Count",
       y = "Malignancy Prevalence") +
  theme_minimal()

ggsave(here("plots/updated_data/cancer_incidence.png"))
```

## Multiplicity adjustment

```{r}
# Store all p-values
p_values <- c(
  cancer_type_species = 9.999e-05,
  library_layout_species = 9.999e-05,
  seq_reads_species = 2.2e-16,
  seq_type_library_layout = 2.2e-16,
  spearman_prevalence = 0.9459,
  linear_prevalence = 0.9479
)

# BH adjusted p-values
adjusted_p <- p.adjust(p_values, method = "BH")
adjusted_p

```

## Output sample counts by species for phylogenetic tree

```{r}
# Replace spaces with underscores in species names for phylogenetic tree
species_samples$organism_scientific_name <- gsub(" ", "_", species_samples$organism_scientific_name)


species_samples <- species_samples %>%
                      rename("Organism" = organism_scientific_name,
                        "Row Count" = Sample_Counts)
species_samples
```

```{r}
# Output mammal sample counts for phylogenetic tree
write.csv(species_samples, here("output_data/non_mouse_data/updated_data/mammal_sample_counts_5-3-2026.csv"), row.names = FALSE)
```

