---
title: "Species Sex Prediction"
output: html_document
---

```{r}
require(tibble) # dataframe manipulation
require(readr) # for read_csv()
require(dplyr) # for mutate()
require(tidyr) # for unnest()
require(purrr) # for map(), reduce()
require(ggplot2)
library(here) # for paths
```

## Read in Enhydra lutris expression data
```{r}
# Read in Enhydra lutris and drop the last 5 rows
enhydra_expr <- read_tsv("/orange/kgraim/panmammalian/data/Enhydra_lutris/PRJNA388419/paired/freya_results/dexseq_count/PRJNA388419.count")

enhydra_expr <- head(enhydra_expr, -5)
```

```{r}
# Extract gene names
enhydra_expr <- enhydra_expr %>%
  mutate(Genes = sub("^[^.]+\\.([^.]+)\\..*$", "\\1", Genes))

# Remove duplicates by taking the median
enhydra_expr <- enhydra_expr %>%
  group_by(Genes) %>%
  summarise(across(where(is.numeric), \(x) median(x, na.rm = TRUE)), .groups = "drop") # drop here means to not keep the grouping in the result

# Check if any duplicates remain
any(duplicated(enhydra_expr$Genes))

```


## Combine species expression data

```{r}
# Function to find and read all bioproject .count files for a given list of mammals
find_and_read_gene_expression_files <- function(base_dir, mammal_names) {
  
  gene_expression_files <- list()
  count_data_list <- list()
  
  # Loop through each mammal in the list
  for (mammal in mammal_names) {
    # Set the base path for the mammal
    mammal_path <- file.path(base_dir, mammal)
    
    # Check if the mammal folder exists
    if (dir.exists(mammal_path)) {
      # Get all bioproject folders under the mammal folder
      bioprojects <- list.dirs(mammal_path, recursive = FALSE, full.names = TRUE)
      
      # Loop through each bioproject folder
      for (bioproject in bioprojects) {
        # Look for 'paired' or 'single' layout folders under each bioproject
        layout_folders <- list.dirs(bioproject, recursive = FALSE, full.names = TRUE)
        
        # Filter for 'paired' and 'single' layout folders
        layout_folders <- layout_folders[grepl("paired|single", layout_folders)]
        
        # Loop through each layout folder and look for the 'dexseq_count' folder for exp data
        for (layout_folder in layout_folders) {
          freya_results_path <- file.path(layout_folder, "freya_results", "dexseq_count")
          
          # Check if the dexseq_count folder exists
          if (dir.exists(freya_results_path)) {
            # Construct the path to the .count file
            count_file <- file.path(freya_results_path, paste0(basename(bioproject), ".count"))
            
            # If the .count file exists, add it to the list of file paths
            if (file.exists(count_file)) {
              gene_expression_files[[length(gene_expression_files) + 1]] <- count_file
            }
          }
        }
      }
    }
  }
  
  # Read all .count files into a list of data frames
  for (count_file in gene_expression_files) {
    count_data <- read_tsv(count_file, col_types = cols(.default = "c"))  
    count_data_list[[length(count_data_list) + 1]] <- count_data
  }

  # Return the list of file paths
  return(list(
    file_paths = gene_expression_files
    ))
}

```

```{r}
# Set directory for mammal data
base_dir <- "/orange/kgraim/panmammalian/data"

# Folders to search for count files
mammal_names <- c(
  "Rattus_norvegicus",
  "Canis_lupus_familiaris",
  "Canis_lupus_familiaris_other_cancers",
  "Mesocricetus_auratus",
  "Chlorocebus_sabaeus",
  "Heterocephalus_glaber",
  "Sus_scrofa",
  "Equus_caballus",
  "Felis_catus",
  "Tupaia_chinensis",
  "Macaca_mulatta",
  "Bos_taurus",
  "Oryctolagus_cuniculus",
  "Bos_indicus",
  "Bubalus_bubalis",
  "Pan_troglodytes",
  "Cricetulus_griseus"
)

result <- find_and_read_gene_expression_files(base_dir, mammal_names)
file_paths <- result$file_paths
```

```{r}
# Load count files and drop the last 5 rows
species_expr <- lapply(file_paths, function(file) {
  data <- readr::read_tsv(file)  
  head(data, -5)                 
})
```

```{r}
# List to store gene counts for each species
gene_counts <- list()

# Iterate through each species to process separately
for (species in mammal_names) {
  species_files <- species_expr[grepl(species, file_paths)]
  if (length(species_files) > 0) {
      # Join the files 
      species_expr_data <- purrr::reduce(species_files, inner_join, by = "Genes") %>%
      # Drop rows with NA in the "Genes" column only
      filter(!is.na(Genes)) %>%
      
      # Get the gene, change gene name to upper case, remove any +- and keep everything after the 2nd underscore
      mutate(Genes = toupper(gsub("[+-]", "", sub("^[^_]*_[^_]*_(.*)", "\\1", Genes)))) %>%
        
      # Remove duplicates by taking median
      dplyr::group_by(Genes) %>%
        summarise(across(where(is.numeric), \(x) median(x, na.rm = TRUE)), .groups = "drop")

    # store current species counts
    gene_counts[[species]] <- species_expr_data
  }
}

```

```{r}
# Add Enhydra lutris into the list
gene_counts[["Enhydra"]] <- enhydra_expr

# Find the number of unique samples for each species
total_unique_samples <- sapply(gene_counts, function(samples) length(unique(samples))-1)
total_unique_samples

# Join gene expression data for all species keeping all genes for sex prediction
all_species_exp <- purrr::reduce(gene_counts, full_join, by = "Genes")
```


```{r}
# Check if any samples (cols) end in '.x' or '.y'
any(grepl("\\.(x|y)$", names(all_species_exp)))
```


```{r}
# There are duplicate samples (between Canis_lupus_familiaris and Canis_lupus_familiaris_other_cancers it seems), so when the data is joined, it renames the samples as sample1.x and sample1.y

# Drop all columns that end with .y
all_species_exp <- all_species_exp %>% dplyr::select(-ends_with(".y"))

# Replace ".x" at the end of column names with nothing
colnames(all_species_exp) <- gsub("\\.x$", "", colnames(all_species_exp))

# Check if any samples (cols) end in '.x' or '.y'
any(grepl("\\.(x|y)$", names(all_species_exp)))

# Check if any duplicate samples (cols)
any(duplicated(names(all_species_exp)))

```


```{r}
# Keep only the first occurrence of each column name
all_species_exp <- all_species_exp[, !duplicated(colnames(all_species_exp))]

# Check
any(duplicated(names(all_species_exp)))
```


```{r}
saveRDS(all_species_exp, here("output_data/non_mouse_data/updated_data/full_join_species_exp_data.rds"))
```


## Load in full join species exp data, if needed
```{r}
all_species_exp <- readRDS(here("output_data/non_mouse_data/updated_data/full_join_species_exp_data.RDS"))

head(all_species_exp)
```

```{r}
# Read in metadata
species_metadata <- readRDS(here("output_data/non_mouse_data/updated_data/species_metadata_harmonized_5-2-26.rds"))

```

```{r}
# Filter exp data for samples that are in the metadata
all_species_exp_filt <- all_species_exp %>%
  dplyr::select(Genes, any_of(species_metadata$run_accession))

```

```{r}
# Rename LOC102156855 to XIST in dog samples
# gene_counts[["Canis_lupus_familiaris"]] <- gene_counts[["Canis_lupus_familiaris"]] %>%
#   mutate(Genes = ifelse(Genes == "LOC102156855", "XIST", Genes)) %>%
#   group_by(Genes) %>%
#   summarise(across(where(is.numeric), median, na.rm = TRUE), .groups = "drop")
# 
# gene_counts[["Canis_lupus_familiaris_other_cancers"]] <- gene_counts[["Canis_lupus_familiaris_other_cancers"]] %>%
#   mutate(Genes = ifelse(Genes == "LOC102156855", "XIST", Genes)) %>%
#   group_by(Genes) %>%
#   summarise(across(where(is.numeric), median, na.rm = TRUE), .groups = "drop")

```

## Rename LOC102156855 to XIST in Canis lupus familiaris samples

```{r}
clf_gene <- "LOC102156855"

# Filter all_species_exp for XIST and LOC102156855
xist_rows_clf <- all_species_exp_filt %>%
  filter(Genes %in% c("XIST", clf_gene))

# Rename LOC102156855 to XIST
xist_rows_clf <- xist_rows_clf %>%
  mutate(Genes = "XIST")

# Collapse these rows into 1
xist_merged_clf <- xist_rows_clf %>%
  group_by(Genes) %>%
  summarise(across(where(is.numeric), \(x) median(x, na.rm = TRUE)), .groups = "drop")

# Remove old XIST and LOC102156855 rows and add the newly collapsed XIST row
all_species_exp_adj <- all_species_exp_filt %>%
  filter(!Genes %in% c("XIST", clf_gene)) %>% # keep all rows except XIST and LOC102156855
  bind_rows(xist_merged_clf)
```

## Rename LOC10699524 to XIST in Macaca mulatta samples

```{r}
mm_gene <- "LOC106995245"

# Filter all_species_exp for XIST and LOC106995245
xist_rows_mm <- all_species_exp_adj %>%
  filter(Genes %in% c("XIST", mm_gene))

# Rename LOC106995245 to XIST
xist_rows_mm <- xist_rows_mm %>%
  mutate(Genes = "XIST")

# Collapse these rows into 1
xist_merged_mm <- xist_rows_mm %>%
  group_by(Genes) %>%
  summarise(across(where(is.numeric), \(x) median(x, na.rm = TRUE)), .groups = "drop")

# Remove old XIST and LOC106995245 rows and add the newly collapsed XIST row
all_species_exp_adj <- all_species_exp_adj %>%
  filter(!Genes %in% c("XIST", mm_gene)) %>% # keep all rows except XIST and LOC106995245
  bind_rows(xist_merged_mm)
```


## Normalization (TMM + log2CPM)

```{r}
library(edgeR)

# function to normalize RNA-seq data
normalize_rnaseq_tmm <- function(counts_df) {
  counts_df_rem <- counts_df %>% column_to_rownames("Genes")
  
  # create a DGEList from the counts dataframe
  expr <- DGEList(counts = counts_df_rem)
  
  # calculate TMM normalization factors
  expr <- calcNormFactors(expr)
  
  # get the normalized counts in log scale (log2CPM) and adds 1
  exp_norm <- as.data.frame(cpm(expr, log = TRUE, prior.count=1))
  
  # return the normalized data frame
  return(exp_norm)
}

```

```{r}
# Fill NAs with zeros to be normalized
species_exp_no_na <- all_species_exp_adj %>% column_to_rownames("Genes")
species_exp_no_na[is.na(species_exp_no_na)] <- 0

# Check for NAs
sum(is.na(species_exp_no_na))
```

```{r}
# Check colsums because of warning when trying to normalize
col_sums <- colSums(species_exp_no_na, na.rm = TRUE)

summary(col_sums)
```

```{r}
# Filter out samples with < 1000 gene expression
species_exp_no_na <- species_exp_no_na[, col_sums >= 1000]

# Check
min(colSums(species_exp_no_na))
```

```{r}
# Normalize exp data using TMM + log2cpm
species_exp_norm <- normalize_rnaseq_tmm(species_exp_no_na %>% rownames_to_column("Genes"))

```

```{r}
# Make a Genes col
species_exp_norm <- species_exp_norm %>% rownames_to_column("Genes")
```


## Key genes for sex prediction

```{r}
# Key genes for determining sex
y_genes <- c("DUSP9", "CD99", "THOC2", "SMARCA1", "TMSB4Y", "RPS4Y1", "DDX3Y", "EIF1AY", "KDM5D", "UTY", "USP9Y", "ZFY")

x_gene <- "XIST"

```

## Create a XIST df
```{r}
x_df <- species_exp_norm %>%
  # Only keep the XIST gene
  filter(Genes == x_gene) %>% 
  
  # Create a long table format where rows are: sample, XIST expression
  pivot_longer(-Genes, names_to = "Sample", values_to = "X") %>%
  
  # Keep everything except the Genes col
  select(-Genes) 

x_df
```

## Create a Y genes df with the sum of Y genes expression

```{r}
y_df <- species_exp_norm %>%
  # Only keep the Y genes
  filter(Genes %in% y_genes) %>%
  
  # Create a long table format so rows have samples and sum of Y genes expression
  pivot_longer(-Genes, names_to = "Sample", values_to = "expression") %>%
  
  # Make expression numeric type
  mutate(expression = as.numeric(expression)) %>%
  
  # Group Y genes by sample ID and sum; set to NA if all NA
  group_by(Sample) %>%
  summarise(
    Y = if (all(is.na(expression))) NA_real_ else sum(expression, na.rm = TRUE),
    .groups = "drop" # remove grouping after
  )

y_df
```

```{r}
# Join the 2 tables by sample ID
plot_data <- left_join(x_df, y_df, by = "Sample")

# Check
plot_data
```

## Scatter plot of current paipu_sex labels

```{r}
# Join current sex labels (metadata) with X and Y expression
plot_labeled <- plot_data %>%
  left_join(
    species_metadata %>% select(run_accession, paipu_sex),
    by = c("Sample" = "run_accession")
  )

# Scatter plot
ggplot(plot_labeled, aes(x = Y, y = X, color = paipu_sex)) +
  geom_point(alpha = 0.7) +
  theme_classic() +
  labs(
    title = "Sex marker expression with known labels",
    x = "Y genes expression",
    y = "X expression",
    color = "paipu_sex labels"
  )

```

## Create a validation df with only known sex labels

```{r}
validation_data <- species_metadata %>%
  # Only keep known labels
  filter(paipu_sex != "UNKNOWN") %>% 
  
  # Sample ID, sex label
  select(run_accession, paipu_sex) %>%
  
  # Join X and Y gene expression with known labels
  left_join(plot_data, by = c("run_accession" = "Sample"))

validation_data
```


```{r}
# Check Y genes expression
validation_data %>%
  group_by(paipu_sex) %>%
  summarise(
    Y_mean = mean(Y, na.rm = TRUE),
    Y_median = median(Y, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

boxplot(Y ~ paipu_sex, data = validation_data, main = "Y genes expression in known sex labels") # Y seems noisy
```

## Check AUC for XIST and Y genes separately and combined

```{r}
library(pROC) # for ROC and AUC

```

```{r}
# Make sex a binary label for ROC with females as the positive class
validation_data <- validation_data %>%
  mutate(sex_binary = ifelse(paipu_sex == "FEMALE", 1, 0))

head(validation_data)
```

```{r}
table(validation_data$paipu_sex)
```

```{r}
# Calculate ROC using Y genes as the predictor to see how well it separates males and females
y_roc <- roc(validation_data$sex_binary, validation_data$Y)

# Calculate AUC to see performance (0.5 means random, 1 mean perf)
auc(y_roc)

# Check the direction of the ROC curve to see how Y behaves in females (positive class)
# If the direction is:
# >: higher Y -> more likely FEMALE (unexpected)
# <: lower Y -> more likely FEMALE (expected)
y_roc$direction

```

```{r}
# Calculate ROC using XIST as the predictor to see how well it separates males and females
x_roc <- roc(validation_data$sex_binary, validation_data$X)

# Calculate AUC to see performance (0.5 means random, 1 mean perf)
auc(x_roc)

# Check the direction of the ROC curve to see how XIST behaves in females (positive class)
# If the direction is:
# >: higher XIST -> more likely FEMALE (expected)
# <: lower XIST -> more likely FEMALE (unexpected, but ROC could flip the direction to keep AUC > 0.5 in pROC)
x_roc$direction
```

```{r}
# Logistic regression model with the combined XIST and Y genes as predictors
model <- glm(sex_binary ~ Y + X, data = validation_data, family = binomial)

# Get the predicted probabilities of females
pred <- predict(model, type = "response")

# Calculate ROC on combined XIST and Y genes predictions
roc_combined <- roc(validation_data$sex_binary, pred)

# Calculate AUC to see performance (0.5 means random, 1 mean perf)
auc(roc_combined)

```

## Predict and label known sex samples

```{r}
# Add predicted probabilities from the model to the validation data
validation_data_pred <- validation_data %>%
  mutate(
    pred_prob = predict(model, type = "response")
  )

# Create a paipu_sex_pred col where if prob > 0.5 classify as female, otherwise male
validation_data_pred <- validation_data_pred %>%
  mutate(
    paipu_sex_pred = ifelse(pred_prob > 0.5, "FEMALE", "MALE")
  )

head(validation_data_pred)
```

## Combined model metrics

```{r}
# Contingency table (true x pred)
cont_table <- table(validation_data_pred$paipu_sex, validation_data_pred$paipu_sex_pred)

cont_table
```

```{r}
# Get confusion matrix vals
TP <- cont_table["FEMALE", "FEMALE"]
TN <- cont_table["MALE", "MALE"]
FP <- cont_table["MALE", "FEMALE"]
FN <- cont_table["FEMALE", "MALE"]

# Total number of samples
total <- sum(cont_table)

# Calculate performance metrics:
# accuracy
accuracy <- (TP + TN) / total
print(paste0("Accuracy: ", round(accuracy * 100, 2), "%"))

# sensitivity (accuracy for females)
sensitivity <- TP / sum(cont_table["FEMALE", ])
print(paste0("Sensitivity: ", round(sensitivity * 100, 2), "%"))

# specificity (accuracy for (males)
specificity <- TN / sum(cont_table["MALE", ])
print(paste0("Specificity: ", round(specificity * 100, 2), "%"))

# precision (female)
precision <- TP / (TP + FP)
print(paste0("Precision: ", round(precision * 100, 2), "%"))

# balanced accuracy
balanced_acc <- (sensitivity + specificity) / 2
print(paste0("Balanced accuracy: ", round(balanced_acc * 100, 2), "%"))

```

## Species specific performance metrics

```{r}
# Join species names with the results
validation_data_pred <- validation_data_pred %>%
  left_join(
    species_metadata %>% select(run_accession, organism_scientific_name),
    by = c("run_accession")
  )

# Calculate performance by species
species_performance <- validation_data_pred %>%
  group_by(organism_scientific_name) %>%
  summarise(
    n = n(), # total number of samples for each species
    accuracy = mean(paipu_sex == paipu_sex_pred),
    sensitivity = mean(paipu_sex_pred[paipu_sex == "FEMALE"] == "FEMALE"),
    specificity = mean(paipu_sex_pred[paipu_sex == "MALE"] == "MALE"),
    .groups = "drop"
  ) %>%
  arrange(accuracy) # sort by accuracy

species_performance
```

## Predict sex for all samples

```{r}
# Predict female probability for all samples 
plot_data_pred <- plot_data %>%
  mutate(
    pred_prob = predict(model, newdata = plot_data, type = "response")
  )

# Create a paipu_sex_pred col where if prob > 0.5 classify as female, otherwise male
plot_data_pred <- plot_data_pred %>%
  mutate(
    paipu_sex_pred = ifelse(pred_prob > 0.5, "FEMALE", "MALE")
  )

plot_data_pred
```

```{r}
# Samples with UNKNOWN sex labels
unknown_pred <- species_metadata %>%
  # Only keep the UNKNOWN sex samples
  filter(paipu_sex == "UNKNOWN") %>%
  select(run_accession, paipu_sex, organism_scientific_name) %>%
  left_join(
    plot_data_pred %>% select(Sample, X, Y, pred_prob, paipu_sex_pred),
    by = c("run_accession" = "Sample")
  )

# Check what predictions were made for the unknown samples
table(unknown_pred$paipu_sex_pred)
```

## Add predictions into the metadata

```{r}
# Rename pred_prob to prob_female for when it gets added to the metadata
plot_data_pred <- plot_data_pred %>%
  dplyr::rename(prob_female = pred_prob)


head(plot_data_pred)
```

```{r}
species_metadata_final <- species_metadata %>%
  left_join(
    plot_data_pred %>% select(Sample, prob_female, paipu_sex_pred),
    by = c("run_accession" = "Sample")
  ) %>%
   mutate(
    # If sex is unknown, fill with predicted value, otherwise keep the original label
    paipu_sex_final = case_when(
      paipu_sex != "UNKNOWN" ~ paipu_sex,
      paipu_sex == "UNKNOWN" ~ paipu_sex_pred
    ),
    paipu_sex_confidence = case_when(
      paipu_sex != "UNKNOWN" ~ "known label",
      prob_female >= 0.9 | prob_female <= 0.1 ~ "high confidence",
      prob_female >= 0.75 | prob_female <= 0.25 ~ "medium confidence",
      TRUE ~ "low confidence" # otherwise low confidence
    )
  ) %>% # Put all new cols right after paipu_sex in the metadata
  relocate(paipu_sex_pred, .after = paipu_sex) %>%
  relocate(prob_female, .after = paipu_sex_pred) %>%
  relocate(paipu_sex_final, .after = prob_female) %>%
  relocate(paipu_sex_confidence, .after = paipu_sex_final)

head(species_metadata_final)
```

```{r}
# Check to make sure all samples have a value in paipu_sex_final
table(species_metadata_final$paipu_sex_final)

```

## Output new metadata

```{r}
# Duplicate metadata
saveRDS(species_metadata_final, file = here("output_data/non_mouse_data/updated_data/species_metadata_harmonized_5-3-26.rds"))

write_tsv(species_metadata_final, file = here("output_data/non_mouse_data/updated_data/species_metadata_harmonized_5-3-26.tsv"))

```

```{r}
# Cols to drop for final metadata
paipu_cols_to_drop <- c("paipu_tissue", "paipu_cell_type_multiple_sources", "paipu_tissue_multiple_sources", "paipu_tissue_explicit", "paipu_tissue_inferred", "paipu_cancer_type_inferred", "paipu_treatment_multiple_sources", "manual_curation_note", "paipu_sex", "paipu_sex_pred", "prob_female")

# Drop unnecessary cols for the final metadata
metadata_harmonized_final <- species_metadata_final %>%
  select(-all_of(paipu_cols_to_drop))

# Final duplicate metadata
saveRDS(metadata_harmonized_final, file = here("output_data/non_mouse_data/updated_data/final_metadata/species_final_dup_metadata_5-3-26.rds"))

write_tsv(metadata_harmonized_final, file = here("output_data/non_mouse_data/updated_data/final_metadata/species_final_dup_metadata_5-3-26.tsv"))

```

```{r}
# Read in deduplicated metadata
species_metadata_no_dup <- readRDS(here("output_data/non_mouse_data/updated_data/species_no_dup_metadata_harmonized_5-2-26.rds"))

# Filter species_metadata_final to only keep samples in species_metadata_no_dup
species_metadata_no_dup_filt <- species_metadata_final %>%
  filter(run_accession %in% species_metadata_no_dup$run_accession)

```

```{r}
# Check
nrow(species_metadata_no_dup)
nrow(species_metadata_no_dup_filt)

```

```{r}
# Check how much in paipu_sex_final for the deduplicated data
table(species_metadata_no_dup_filt$paipu_sex_final)

```

```{r}
# Deduplicated metadata
saveRDS(species_metadata_no_dup_filt, file = here("output_data/non_mouse_data/updated_data/species_no_dup_metadata_harmonized_5-3-26.rds"))

write_tsv(species_metadata_no_dup_filt, file = here("output_data/non_mouse_data/updated_data/species_no_dup_metadata_harmonized_5-3-26.tsv"))

```

```{r}
# Drop unnecessary cols for the final metadata
metadata_harmonized_no_dup_final <- species_metadata_no_dup_filt %>%
  select(-all_of(paipu_cols_to_drop))

# Final deduplicated metadata
saveRDS(metadata_harmonized_no_dup_final, file = here("output_data/non_mouse_data/updated_data/final_metadata/species_final_no_dup_metadata_5-3-26.rds"))

write_tsv(metadata_harmonized_no_dup_final, file = here("output_data/non_mouse_data/updated_data/final_metadata/species_final_no_dup_metadata_5-3-26.tsv"))
```

