---
title: "Species Normalization, Duplicate Removal and PCA"
output: html_notebook
---
```{r}
require(tibble) # dataframe manipulation
require(readr) # for read_csv()
require(dplyr) # for mutate()
require(tidyr) # for unnest()
require(purrr) # for map(), reduce()
require(edgeR) # for normalization
library(here) # for paths
```

```{r}
# Read in harmonized metadata
all_metadata_orig <- readRDS(here("output_data/non_mouse_data/updated_data/species_metadata_harmonized_5-2-26.rds"))

# Rename some of the columns
all_metadata <- all_metadata_orig %>%
  rename_with(~c("Sample", "BioProject", "Species", "CancerType", "CancerTypeFinal", "Tissue", "CancerSystem"),
              .cols = c("run_accession", "bioproject", "organism_scientific_name", "paipu_cancer_type", "paipu_cancer_type_final", "paipu_tissue_final", "paipu_cancer_system")) 

# Only keep certain columns in the metadata
all_metadata <- all_metadata %>%
  dplyr::select(Sample, BioProject, Species, CancerType, CancerTypeFinal, Tissue, CancerSystem)

# Check
head(all_metadata)

```

```{r}
# Read in species raw exp data
all_species_exp <- readRDS(here("output_data/non_mouse_data/updated_data/species_exp_raw.rds"))

head(all_species_exp)
```


```{r}
# Only keep exp data for the samples in the metadata
all_species_exp <- all_species_exp %>%
  dplyr::select(1, # keep the gene col
    dplyr::any_of(all_metadata$Sample))

# Check if metadata and exp data are in the same order and have the same samples
all.equal(unname(all_metadata$Sample), unname(colnames(all_species_exp)[-1])) # use for .rds
```


```{r}
# Make the exp data be in the same order as the metadata
all_species_exp <- all_species_exp %>%
  dplyr::select("Genes", all_metadata$Sample)

# Check
all.equal(unname(all_metadata$Sample), unname(colnames(all_species_exp)[-1])) # use for .rds
```

```{r}
head(all_species_exp)
```
```{r}
unique(all_metadata$Species)
```


## Species Colors and PCA function

```{r}
# New species colors based on taxonomic group
species_colors <- c(
  # Carnivora
  "Canis lupus familiaris" = "#C02020",
  "Enhydra lutris" = "#CD4D4D",
  "Felis catus" = "#861616",
  
  # Cetartiodactyla
  "Bos indicus" = "#6C0BCE",
  "Bos taurus" = "#6E4895",
  "Bubalus bubalis" = "#D4AEFB",
  "Sus scrofa" = "#B878F8",
  
  # Lagomorpha
  "Oryctolagus cuniculus" = "#DA1884",
  
  # Perissodactyla
  "Equus caballus" = "#FFA500",
  
  # Primate
  "Chlorocebus sabaeus" = "#00664D",
  "Gorilla gorilla" = "#003325",
  "Macaca fascicularis" = "#00D49B",
  "Macaca mulatta" = "#008000",
  "Pan troglodytes" = "#66C5AB",
  
  # Rodentia
  "Cricetulus griseus" = "#6E7CD1",
  "Mesocricetus auratus" = "#2E3C90",
  "Heterocephalus glaber" = "#4C5BB5",
  "Rattus norvegicus" = "#9EA8E0",

  
  # Scandentia
  "Tupaia chinensis" = "#4DBBD5"
)
```

```{r}
library(ggplot2)

plot_folder <- "plots/updated_data"

visualize_pca <- function(exp, met, title, sp_colors){
  # set Genes as row names
  all_exp_filtered_rem <- exp %>% column_to_rownames("Genes")
  
  # transpose the data to make the genes in columns and samples in rows
  exp_transposed <- t(all_exp_filtered_rem)
  
  # remove 0 values to plot PCA
  exp_subset <- exp_transposed[, which(apply(exp_transposed, 2, var) != 0), drop = FALSE]
  
  # perform PCA
  df_pca <- prcomp(exp_subset, scale = TRUE, center = TRUE)
  
  # add species metadata
  df <- data.frame(
    PC1 = df_pca$x[,1],
    PC2 = df_pca$x[,2],
    Species = met$Species
  )
  
  # plot pca
  pca_plot <- ggplot(df, aes(x = PC1, y = PC2, color = Species)) +
    geom_point(size = 2, alpha = 0.8) + # size use to be 3
    labs(x = "PC1", y = "PC2") +
    scale_color_manual(values = sp_colors) + 
    theme_minimal() + # white background
    theme(
      axis.line = element_line(colour = "black"), # bold axis lines
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA)
    )
  
  # save pca plot
  ggsave(
    filename = paste0(title, ".pdf"),
    plot = pca_plot,
    path = here(plot_folder),
    width = 10, height = 8,
    #limitsize = FALSE,
    device = cairo_pdf
  )
  
  return(pca_plot)
}

```

```{r}
# Plot PCA
raw_pca <- visualize_pca(all_species_exp, all_metadata, "Species_Raw_PCA_5-2-26", species_colors)

```

```{r}
# Check for NA 
sum(is.na(all_species_exp[-1]))

# Check if there are duplicate genes
sum(duplicated(all_species_exp$Genes))
```

```{r}
# Set Genes as row names
all_species_exp_adj <- all_species_exp %>% column_to_rownames("Genes")

# Calculate sample expression sums
samples_total_exp <- colSums(all_species_exp_adj)

```

```{r}
# Number of samples in the dataset
print(paste0("Number of samples in dataset: ", ncol(all_species_exp_adj)))

# Number of samples where total expression > 0
print(paste0("Number of samples where total expression > 0: ", sum(samples_total_exp > 0)))

```

```{r}
# Get samples whose total expression is zero
zero_samples <- names(samples_total_exp[samples_total_exp == 0])

# Filter out samples where total expression is zero
all_species_exp_adj <- all_species_exp_adj[, !(colnames(all_species_exp_adj) %in% zero_samples)]

# Filter metadata to match
all_metadata <- all_metadata %>%
  filter(Sample %in% colnames(all_species_exp_adj))

# Check if metadata and exp in the same order
all.equal(unname(all_metadata$Sample), unname(colnames(all_species_exp_adj)))
```

```{r}
# Number of genes in the dataset
print(paste0("Number of genes in dataset: ", nrow(all_species_exp_adj)))

# Number of genes where the sum of expression across samples > 0
print(paste0("Number of genes where total expression > 0: ", sum(rowSums(all_species_exp_adj) > 0)))

```

# Filter low quality samples

```{r}
# Calculate num of samples with non-zero exp in each sample
nonzero <- colSums(all_species_exp_adj > 0)

# Check the distribution of non-zero genes
summary(nonzero)
```

```{r}
# Plot histogram of number of expressed genes per sample with 50 bins
hist_plot <- hist(nonzero, breaks = 50, main = "Number of expressed genes per sample", xlab = "Count")

```


```{r}
# Save hist boundaries
hist_boundaries <- data.frame(
  bin_left = hist_plot$breaks[-length(hist_plot$breaks)],
  bin_right = hist_plot$breaks[-1],
  count = hist_plot$counts
)
```

```{r}
# Check the low count area (< 4000)
subset_bins <- hist_boundaries[hist_boundaries$bin_right <= 4000, ]
subset_bins
```

```{r}
# Compute differences
subset_bins$diff <- c(NA, diff(subset_bins$count))

# Find the knee in this region
knee_bin <- subset_bins[which.max(subset_bins$diff), ]

knee_bin
```

```{r}
# Set the cutoff as the left of the knee bin
cutoff <- knee_bin$bin_left
```

```{r}
# Save plot as pdf
pdf(here("plots/updated_data/Gene_Expression_Sample_Filtering.pdf"), width = 7, height = 5)

# Plot a new histogram with cutoff line
hist_plot <- hist(
  nonzero,
  breaks = 50,
  main = "Non-zero genes per sample",
  xlab = "Count",
  col = "steelblue"
)

# Create red line for cutoff on histogram
abline(v = cutoff, col = "red", lwd = 2, lty = 2)

# Add a cutoff label
text(
  x = cutoff,
  y = max(hist_plot$counts) * 0.9, # put the text a little higher than the tallest bar
  labels = paste("cutoff =", cutoff),
  col = "red",
  pos = 2 # place text to the left of the point (right = 4)
)

dev.off()
```

```{r}
# Number of samples before filtering
ncol(all_species_exp_adj)
```

```{r}
# Only keep samples above the cutoff
keep_samples <- nonzero > cutoff

# Filter exp data
all_species_exp_adj_filtered <- all_species_exp_adj[, keep_samples]

# Number of samples after filtering
ncol(all_species_exp_adj_filtered)
```

```{r}
# Filter metadata to only keep samples in the filtered exp
all_metadata_filtered <- all_metadata[
  all_metadata$Sample %in% colnames(all_species_exp_adj_filtered),
]

# Make exp and metadata be in the same order
all_metadata_filtered <- all_metadata_filtered[
  match(colnames(all_species_exp_adj_filtered), all_metadata_filtered$Sample),
]

# Check
all(colnames(all_species_exp_adj_filtered) == all_metadata_filtered$Sample)
```

## Filter lowly expressed genes

```{r}
library(edgeR)

# Create an edgeR object
dge <- DGEList(counts = all_species_exp_adj_filtered)

# Filter lowly expressed genes preserving expression in Species and CancerType
keep_genes <- filterByExpr(dge, group = interaction(all_metadata_filtered$Species, all_metadata_filtered$CancerType))

# Filter exp to only keep certain genes
all_species_exp_adj_filtered <- all_species_exp_adj_filtered[keep_genes, ]

```

## Normalization (TMM + log2CPM)

```{r}
#library(DESeq2) 

# function to normalize RNA-seq data
normalize_rnaseq_tmm <- function(counts_df) {
  counts_df_rem <- counts_df %>% column_to_rownames("Genes")
  
  # create a DGEList from the counts dataframe
  expr <- DGEList(counts = counts_df_rem)
  
  # calculate TMM normalization factors
  expr <- calcNormFactors(expr)
  
  # get the normalized counts in log scale (log2CPM) and adds 1
  exp_norm <- as.data.frame(cpm(expr, log = TRUE, prior.count=1))
  
  # return the normalized data frame
  return(exp_norm)
}

```

```{r}
# Normalize exp data using TMM + log2cpm
species_exp_norm <- normalize_rnaseq_tmm(all_species_exp_adj_filtered %>% rownames_to_column("Genes"))

```

## PCA of normalized data
```{r}
norm_pca <- visualize_pca(species_exp_norm %>% rownames_to_column("Genes"), all_metadata_filtered, "Normalized_Species_PCA_with_dup_5-2-26", species_colors)
```

# Pairwise sample correlation

```{r}
# Check if the order matches between expr and metadata
all(all_metadata_filtered$Sample == colnames(species_exp_norm))

```

```{r}
head(species_exp_norm)
```

```{r}
# Check if the data is numeric
is.numeric(as.matrix(species_exp_norm))
```

```{r}
# List to store sample pairs that have high correlation per species
deduplicate_pairs_by_species <- list()
cor_by_species <- c()

for (species_name in unique(all_metadata_filtered$Species)) {
  # Store sample indices for current species
  species_sample_idx <- which(all_metadata_filtered$Species == species_name)

  # Skip species with < 2 samples
  if (length(species_sample_idx) < 2) next

  # Subset exp data for current species
  subset_species_expr <- species_exp_norm[, species_sample_idx, drop = FALSE]

  # Compute pairwise sample correlations within species
  species_cor_matrix <- cor(subset_species_expr, method = "pearson")

  # Get all upper triangle correlations
  cor_vals <- species_cor_matrix[upper.tri(species_cor_matrix)]
  
  # Store species correlations
  cor_by_species <- c(cor_by_species, cor_vals)
}

```

```{r}
# Plot histogram to choose cutoff
hist_plot2 <- hist(
  cor_by_species,
  breaks = 100,
  main = "Within Species Sample Correlations Distribution",
  xlab = "Pearson correlation",
  col = "steelblue"
)

```

```{r}
# Check bin boundaries to determine cutoff on chart
hist_boundaries2 <- data.frame(
  bin_left = hist_plot2$breaks[-length(hist_plot2$breaks)],
  bin_right = hist_plot2$breaks[-1],
  count = hist_plot2$counts
)

tail(hist_boundaries2) # put cutoff at 0.99
```


```{r}
# Save plot as pdf
pdf(here("plots/updated_data/Correlation_Sample_Filtering.pdf"), width = 7, height = 5)

#par(xpd = NA) # allows text outside of the plot region (the cutoff label was being cut off the chart)

# Plot a new histogram with cutoff line
hist_plot2 <- hist(
  cor_by_species,
  breaks = 100,
  main = "Within Species Sample Correlations Distribution",
  xlab = "Pearson correlation",
  col = "steelblue"
)

# Set cutoff
hist_cutoff <- 0.99

# Create red line for cutoff on histogram
abline(v = hist_cutoff, col = "red", lwd = 2, lty = 2)

# Add a cutoff label
text(
  x = hist_cutoff,
  y = max(hist_plot2$counts) * 0.9, # put the text a little higher than the tallest bar
  labels = paste("cutoff =", hist_cutoff),
  col = "red",
  pos = 2 # place text to the left of the point (right = 4)
)

dev.off()
```


```{r}
# Remove samples that have high correlations with another sample using the cutoff r=0.99 + edit metadata to match. Output 2 datasets, 1 w/ duplicates, 1 w/o duplicates

# Set cutoff
cutoff <- 0.99

# Stores samples to remove from exp data
samples_to_remove <- c()

for (species_name in unique(all_metadata_filtered$Species)) {
  # Store sample indices for current species
  species_sample_idx <- which(all_metadata_filtered$Species == species_name)

  # Skip species with <2 samples
  if (length(species_sample_idx) < 2) next

  # Subset exp data for current species
  subset_species_expr <- species_exp_norm[, species_sample_idx, drop = FALSE]

  # Compute pairwise sample correlations within species
  species_cor_matrix <- cor(subset_species_expr, method = "pearson")

  # Avg correlation per sample for current species
  avg_cor <- rowMeans(species_cor_matrix)
  
  # Identify sample pair indices that are above the cutoff
  high_cor_idx <- which(
    upper.tri(species_cor_matrix) &
    species_cor_matrix > cutoff,
    arr.ind = TRUE
  )

  # If no high correlations, go to the next species
  if (nrow(high_cor_idx) == 0) next
  
  # Iterate through high correlation pairs
  for (i in seq_len(nrow(high_cor_idx))) {
    # Store the indices of the current pair to get the samples
    sample1_idx <- high_cor_idx[i, 1]
    sample2_idx <- high_cor_idx[i, 2]
    
    sample1 <- species_sample_idx[sample1_idx]
    sample2 <- species_sample_idx[sample2_idx]
    
    # If any of the current samples are already in the list to be removed, move on to the next pair
    if (sample1 %in% samples_to_remove || sample2 %in% samples_to_remove) next
    
    # Otherwise, remove the sample with the highest avg correlation
    if (avg_cor[sample1_idx] >= avg_cor[sample2_idx]) {
      samples_to_remove <- c(samples_to_remove, sample1)
    } 
    else {
      samples_to_remove <- c(samples_to_remove, sample2)
    }
  }
}

```

```{r}
# Filter out samples stored in samples_to_remove
samples_to_keep <- setdiff(seq_len(ncol(species_exp_norm)), samples_to_remove)
species_exp_no_dup <- species_exp_norm[, samples_to_keep, drop = FALSE]
species_metadata_no_dup <- all_metadata_filtered[samples_to_keep, , drop = FALSE]
```

```{r}
# Check if all correlations are now < 0.99
#max(cor(species_exp_no_dup)[upper.tri(cor(species_exp_no_dup))])
```

```{r}
print("Before removing duplicates: ")
nrow(all_metadata_filtered)
ncol(species_exp_norm)

print("After removing duplicates: ")
nrow(species_metadata_no_dup)
ncol(species_exp_no_dup)
```
```{r}
unique(all_metadata_filtered$Species)
```


## PCA normalized with duplicates removed

```{r}
# Normalized PCA without duplicates
norm_pca_no_dup <- visualize_pca(species_exp_no_dup %>% rownames_to_column("Genes"), species_metadata_no_dup, "Normalized_Species_PCA_no_dup_5-2-26", species_colors)

```


## Output simplified and full metadata and exp data (duplicated)

```{r}
# Filter full metadata to only keep duplicated high quality samples
full_metadata_dup <- all_metadata_orig[
  match(all_metadata_filtered$Sample, all_metadata_orig$run_accession),
]

# Check that the full and simplified metadata have the same samples
all(full_metadata_dup$run_accession == all_metadata_filtered$Sample)

# Check if the order matches between expr and full metadata
all(full_metadata_dup$run_accession == colnames(species_exp_norm))

# Check if the order matches between expr and simplified metadata
all(all_metadata_filtered$Sample == colnames(species_exp_norm))

```


```{r}
# Save simplified metadata duplicated
saveRDS(all_metadata_filtered, file = here("output_data/non_mouse_data/updated_data/simplified_species_metadata_5-2-26.rds"))

#write.csv(all_metadata_filtered, file = here("output_data/non_mouse_data/updated_data/simplified_species_metadata_5-2-26.csv"), row.names = FALSE)

write_tsv(all_metadata_filtered, file = here("output_data/non_mouse_data/updated_data/simplified_species_metadata_5-2-26.tsv"))

# Save full metadata duplicated
saveRDS(full_metadata_dup, file = here("output_data/non_mouse_data/updated_data/species_metadata_harmonized_5-2-26.rds"))

#write.csv(full_metadata_dup, file = here("output_data/non_mouse_data/updated_data/species_metadata_harmonized_5-2-26.csv"), row.names = FALSE)

write_tsv(full_metadata_dup, file = here("output_data/non_mouse_data/updated_data/species_metadata_harmonized_5-2-26.tsv"))

# Save duplicated exp data
saveRDS(species_exp_norm %>% rownames_to_column("Genes"), file = here("output_data/non_mouse_data/updated_data/species_exp_tmm_logcpm.rds"))

write_tsv(species_exp_norm %>% rownames_to_column("Genes"), file = here("output_data/non_mouse_data/updated_data/species_exp_tmm_logcpm.tsv"))

```

```{r}
# Cols to drop for final metadata
paipu_cols_to_drop <- c("cancer_type", "paipu_tissue", "paipu_cell_type_multiple_sources", "paipu_tissue_multiple_sources", "paipu_tissue_explicit", "paipu_tissue_inferred", "paipu_cancer_type_inferred", "paipu_treatment_multiple_sources", "manual_curation_note")

metadata_harmonized_final <- full_metadata_dup %>%
  select(-all_of(paipu_cols_to_drop))

## Output FINAL harmonized species metadata (duplicates)
# Save as RDS
saveRDS(metadata_harmonized_final, file = here("output_data/non_mouse_data/updated_data/final_metadata/species_final_dup_metadata_5-2-26.rds"))

# Save as CSV
#write.csv(metadata_harmonized_final, file = here("output_data/non_mouse_data/updated_data/final_metadata/species_final_dup_metadata_5-2-26.csv"), row.names = FALSE)

write_tsv(metadata_harmonized_final, file = here("output_data/non_mouse_data/updated_data/final_metadata/species_final_dup_metadata_5-2-26.tsv"))
```


## Output simplified, full and final metadata and exp data (deduplicated)

```{r}
# Filter full metadata to only keep non-duplicated samples
full_metadata_no_dup <- all_metadata_orig[
  match(species_metadata_no_dup$Sample, all_metadata_orig$run_accession),
]

# Check if full and simplified deduplicated metadata have the same samples
all(full_metadata_no_dup$run_accession == species_metadata_no_dup$Sample)
```

```{r}
# Save simplified metadata non-duplicated
saveRDS(species_metadata_no_dup, file = here("output_data/non_mouse_data/updated_data/simplified_no_dup_species_metadata_5-2-26.rds"))

#write.csv(species_metadata_no_dup, file = here("output_data/non_mouse_data/updated_data/simplified_no_dup_species_metadata_5-2-26.csv"), row.names = FALSE)

write_tsv(species_metadata_no_dup, file = here("output_data/non_mouse_data/updated_data/simplified_no_dup_species_metadata_5-2-26.tsv"))

# Save full metadata non-duplicated
saveRDS(full_metadata_no_dup, file = here("output_data/non_mouse_data/updated_data/species_no_dup_metadata_harmonized_5-2-26.rds"))

#write.csv(full_metadata_no_dup, file = here("output_data/non_mouse_data/updated_data/species_no_dup_metadata_harmonized_5-2-26.csv"), row.names = FALSE)

write_tsv(full_metadata_no_dup, file = here("output_data/non_mouse_data/updated_data/species_no_dup_metadata_harmonized_5-2-26.tsv"))

# Save non-duplicated exp data
saveRDS(species_exp_no_dup %>% rownames_to_column("Genes"), file = here("output_data/non_mouse_data/updated_data/species_no_dup_exp_tmm_logcpm.rds"))

write_tsv(species_exp_no_dup %>% rownames_to_column("Genes"), file = here("output_data/non_mouse_data/updated_data/species_no_dup_exp_tmm_logcpm.tsv"))

```

```{r}
metadata_harmonized_final_no_dup <- full_metadata_no_dup %>%
  select(-all_of(paipu_cols_to_drop))

## Output FINAL harmonized species metadata (no duplicates)
# Save as RDS
saveRDS(metadata_harmonized_final_no_dup, file = here("output_data/non_mouse_data/updated_data/final_metadata/species_final_no_dup_metadata_5-2-26.rds"))

# Save as CSV
#write.csv(metadata_harmonized_final_no_dup, file = here("output_data/non_mouse_data/updated_data/final_metadata/species_final_dup_metadata_5-2-26.csv"), row.names = FALSE)

write_tsv(metadata_harmonized_final_no_dup, file = here("output_data/non_mouse_data/updated_data/final_metadata/species_final_no_dup_metadata_5-2-26.tsv"))
```


## Visualize PCA - colored by cancer system

```{r}
library(Polychrome)
data("glasbey")

# Function to view pca colored by CancerSystem
visualize_pca_cs <- function(exp, met, title, color_by = "CancerSystem", custom_colors = NULL){
  # Set Genes as row names
  all_exp_filtered_rem <- exp %>% column_to_rownames("Genes")
  
  # Transpose the data (genes in columns, samples in rows)
  exp_transposed <- t(all_exp_filtered_rem)
  
  # Remove zero-variance genes
  exp_subset <- exp_transposed[, which(apply(exp_transposed, 2, var) != 0), drop = FALSE]
  
  # Perform PCA
  df_pca <- prcomp(exp_subset, scale = TRUE, center = TRUE)
  
  # flip up/down NEW comment/uncomment based on if chart needs to be flipped
  #df_pca$x[,2] <- -df_pca$x[,2]
  
  # Prepare PCA dataframe
  df <- data.frame(
    PC1 = df_pca$x[,1],
    PC2 = df_pca$x[,2],
    ColorGroup = met[[color_by]],
    Species = if ("Species" %in% colnames(met)) met$Species else NA
  )
  
  # If no custom colors, use viridis
  if (is.null(custom_colors)) {
    groups <- unique(df$ColorGroup)
    n_colors <- length(groups)
    custom_colors <- glasbey[1:n_colors]
    names(custom_colors) <- groups
  }
  
  # Base ggplot setup
  pca_plot <- ggplot(df, aes(x = PC1, y = PC2)) +
    labs(x = "PC1", y = "PC2", color = "Cancer System") +
    theme_minimal() + # white background
    theme(
      axis.line = element_line(colour = "black"), # bold axis lines
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA)
    )
  
  pca_plot <- pca_plot + 
    geom_point(aes(color = ColorGroup), size = 2, alpha = 0.8)

  
  # Apply color scale here (always after geom_point)
  pca_plot <- pca_plot + scale_color_manual(values = custom_colors)
  
  # Save PCA plot as pdf
  ggsave(
    filename = paste0(title, ".pdf"),
    plot = pca_plot,
    path = here(plot_folder),
    width = 10, height = 8,
    #limitsize = FALSE,
    device = cairo_pdf
  )
  
  return(pca_plot)
}

```


```{r}
# Check unique cancer systems
unique(species_metadata_no_dup$CancerSystem)
```

```{r}
cs_colors <- c(
  "Breast" = "#4E79A7",
  "Digestive_Gastrointestinal" = "#F28E2B",
  "Endocrine_Neuroendocrine" = "#E15759",
  "Eye" = "#76B7B2",
  "Genitourinary" = "#59A14F",
  "Gynecologic" = "#EDC948",
  "Head and Neck" = "#B07AA1",
  "Hematologic_Blood" = "#FF9DA7",
  "Musculoskeletal" = "#9C755F",  # updated spelling
  "Neurologic" = "#BAB0AC",
  "Respiratory_Thoracic" = "#1F77B4",
  "Skin" = "#AEC7E8"
  #"Unspecified" = "#17BECF"
)

# Check
cs_colors

```

```{r}
# Edit metadata to label cancersystem without underscores, replace with '/'
met_pca <- species_metadata_no_dup
met_pca$CancerSystem_plot <- gsub("_", "/", met_pca$CancerSystem)


# CancerSystem pca normalized
cs_norm_pca_sp <- visualize_pca_cs(species_exp_no_dup %>% rownames_to_column("Genes"), met_pca, "Normalized_Species_No_Dup_CancerSystem_PCA_5-2-26", color_by = "CancerSystem_plot", custom_colors = setNames(cs_colors, gsub("_", "/", names(cs_colors))))

```

## Output PCAs as a figure in a 1x1 grid 

```{r}
library(patchwork)
```


```{r}
# Create grid with PCA after normalization no duplicates colored by species and colored by cancer system
#pca_fig <- (norm_pca_no_dup + cs_norm_pca) + plot_annotation(tag_levels = "A") &
  #theme(plot.tag = element_text(size = 22)) # increase panel font sizes

pca_fig_sp <- (norm_pca_no_dup + cs_norm_pca_sp) + plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 22), # panel labels size
        legend.text = element_text(size = 16), # legend font size
        legend.title = element_text(size = 16), # legend title size
        legend.key.size = unit(0.6, "cm")) & # legend symbols size
  guides(
    color = guide_legend(override.aes = list(size = 3)) # legend point size
  )

pca_fig_sp
```

```{r}
# Save figure as pdf
ggsave(here(plot_folder, "AtlasPCA.pdf"), pca_fig_sp, width = 20, height = 10, device = cairo_pdf)

```



