---
title: "Species and Treehouse Data PCA"
output: html_document
---

```{r}
library(tidyverse)
library(edgeR)
library(sva)
library(ggplot2)
library(here)
```

## Read in Treehouse and Mammal datasets

```{r}
# Treehouse
treehouse_expr <- readRDS(here("output_data/treehouse_exp_hugo.rds"))
treehouse_metadata <- readRDS(here("output_data/treehouse_metadata.rds"))

# Other mammals normalized data
mammals_expr_norm <- readRDS(here("output_data/non_mouse_data/updated_data/species_no_dup_exp_tmm_logcpm.rds"))

mammals_metadata <- readRDS(here("output_data/non_mouse_data/updated_data/simplified_no_dup_species_metadata_5-2-26.rds"))

```

```{r}
# Make Genes rownames
treehouse_expr <- treehouse_expr %>%
  column_to_rownames("Genes") %>%
  as.matrix()

mammals_expr_norm <- mammals_expr_norm %>%
  column_to_rownames("Genes") %>%
  as.matrix()
```


```{r}
# Check if they're in the same order
all(colnames(treehouse_expr) == treehouse_metadata$Sample)
all(colnames(mammals_expr_norm) == mammals_metadata$Sample)

```


```{r}
# Rename cancer_system to CancerSystem in treehouse metadata
treehouse_metadata <- treehouse_metadata %>%
  dplyr::rename(
    CancerSystem = cancer_system,
    Species = organism
    )
```


```{r}
# Check numeric
is.numeric(treehouse_expr[1, 1])
is.numeric(mammals_expr_norm[1, 1])

# Check for NAs
anyNA(treehouse_expr)
anyNA(mammals_expr_norm)

```

```{r}
# Any all 0 genes?
print("Number of all zero genes: ")
print(paste0("Mammals: ", sum(rowSums(mammals_expr_norm) == 0)))
print(paste0("Treehouse: ", sum(rowSums(treehouse_expr) == 0)))

cat("\n") # empty line

# Any all 0 samples?
print("Number of all zero samples: ")
print(paste0("Mammals: ", sum(colSums(mammals_expr_norm) == 0)))
print(paste0("Treehouse: ", sum(colSums(treehouse_expr) == 0)))

```

```{r}
# Filter out treehouse all zero genes
treehouse_expr <- treehouse_expr[rowSums(treehouse_expr) > 0, ]

# Check 
sum(rowSums(treehouse_expr) == 0)
```


## Visualize PCA

```{r}
# Create pca_group for visualizing the pca
treehouse_metadata$pca_group <- dplyr::case_when(
  treehouse_metadata$age_group == "adult" ~ "Homo sapiens - Adult",
  treehouse_metadata$age_group == "pediatric" ~ "Homo sapiens - Pediatric",
  TRUE ~ NA_character_
)

mammals_metadata$pca_group <- mammals_metadata$Species

```

```{r}
# New species colors based on taxonomic group
species_colors <- c(
  # Carnivora
  "Canis lupus familiaris" = "#C02020",
  "Enhydra lutris" = "#CD4D4D",
  "Felis catus" = "#861616",
  
  # Cetartiodactyla
  "Bos indicus" = "#6C0BCE",
  "Bos taurus" = "#6E4895",
  "Bubalus bubalis" = "#D4AEFB",
  "Sus scrofa" = "#B878F8",
  
  # Lagomorpha
  "Oryctolagus cuniculus" = "#DA1884",
  
  # Perissodactyla
  "Equus caballus" = "#FFA500",
  
  # Primate
  "Chlorocebus sabaeus" = "#00664D",
  "Gorilla gorilla" = "#003325",
  "Macaca fascicularis" = "#00D49B",
  "Macaca mulatta" = "#008000",
  "Pan troglodytes" = "#66C5AB",
 
  # Rodentia
  "Cricetulus griseus" = "#6E7CD1",
  "Mesocricetus auratus" = "#2E3C90",
  "Heterocephalus glaber" = "#4C5BB5",
  "Rattus norvegicus" = "#9EA8E0",
  
  # Scandentia
  "Tupaia chinensis" = "#4DBBD5"
)

# Primates
human_colors <- c(
  "Homo sapiens - Pediatric" = "#A3A500",
  "Homo sapiens - Adult" = "#4D9221"   
)

# Combine the colors
pca_colors <- c(human_colors, species_colors)

```


```{r}
plot_folder <- "plots/updated_data/species+human"

visualize_pca <- function(exp, met, title, group_col, colors, legend_title) {

  # If Genes is a column, make it rownames
  if ("Genes" %in% colnames(exp)) {
    exp <- column_to_rownames(exp, "Genes")
  }

  # Transpose the data to make the genes in columns and samples in rows
  exp_transposed <- t(as.matrix(exp))

  # Remove 0 variance genes to plot PCA
  exp_subset <- exp_transposed[, apply(exp_transposed, 2, var) != 0, drop=FALSE]

  # Perform PCA
  df_pca <- prcomp(exp_subset, center = TRUE, scale. = TRUE)

  # Add metadata and create pca dataframe
  df <- data.frame(
    PC1 = df_pca$x[, 1],
    PC2 = df_pca$x[, 2],
    Group = met[[group_col]]
  )

  # Plot PCA
  pca_plot <- ggplot(df, aes(x = PC1, y = PC2, color = Group)) +
    geom_point(size = 2, alpha = 0.6) +
    labs(x = "PC1", y = "PC2") +
    scale_color_manual(values = colors, name = legend_title) +
    theme_minimal() +
    theme(
      axis.line = element_line(colour = "black"),
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA)
    )
  
  # Save pca plot
  ggsave(
    filename = paste0(title, ".pdf"),
    plot = pca_plot,
    path = here(plot_folder),
    width = 10, height = 8,
    #limitsize = FALSE,
    device = cairo_pdf
  )

  return(pca_plot)
}

```

## PCA of treehouse raw
```{r}
visualize_pca(
  exp = treehouse_expr,
  met = treehouse_metadata,
  title = "Treehouse_Raw_PCA_5-2-26",
  group_col = "pca_group",
  colors = pca_colors,
  legend_title = "Species"
)
```

## TMM Normalize and logCPM

```{r}
# Function to normalize RNA-seq data
normalize_rnaseq_tmm <- function(counts_mat) {
  # Create a DGEList from the counts matrix
  expr <- DGEList(counts = counts_mat)
  
  # Calculate TMM normalization factors
  expr <- calcNormFactors(expr)
  
  # Convert counts to log scale (log2CPM) and adds 1
  exp_norm <- cpm(expr, log = TRUE, prior.count = 1)
  
  # return the normalized data
  return(exp_norm)
}

```

```{r}
# Normalize treehouse
treehouse_expr_norm <- normalize_rnaseq_tmm(treehouse_expr)

```


## PCA of treehouse normalized
```{r}
visualize_pca(
  exp = treehouse_expr_norm,
  met = treehouse_metadata,
  title = "Treehouse_Normalized_PCA_4-16-26",
  group_col = "pca_group",
  colors = pca_colors,
  legend_title = "Species"
)

```

## Combine the mammals norm + treehouse norm exp data
```{r}
# Get common genes in treehouse and mammal normalized data
common_genes_norm <- intersect(rownames(treehouse_expr_norm), rownames(mammals_expr_norm))

length(common_genes_norm)
```

```{r}
# Only keep the common genes
mammals_norm_expr_common  <- mammals_expr_norm[common_genes_norm, , drop = FALSE]

treehouse_norm_expr_common <- treehouse_expr_norm[common_genes_norm, , drop = FALSE]

```

```{r}
# Combine expr data
combined_expr_norm <- cbind(mammals_norm_expr_common, treehouse_norm_expr_common)

```

## Combine human + species metadata
```{r}
# Create a CancerTypeFinal col for the treehouse metadata
treehouse_metadata$CancerTypeFinal <- treehouse_metadata$CancerType

head(treehouse_metadata)
```

```{r}
combined_metadata_norm <- bind_rows(mammals_metadata, treehouse_metadata)
```

```{r}
# Reorder metadata to match the expr data
combined_metadata_norm <- combined_metadata_norm[
  match(colnames(combined_expr_norm), combined_metadata_norm$Sample),
]

# Check
all(colnames(combined_expr_norm) == combined_metadata_norm$Sample)

```

```{r}
# Check
print("Combined metadata and expr data dimensions: ")
dim(combined_metadata_norm)
dim(combined_expr_norm)

```

### ComBat batch correction human vs non-human w/ CancerType preserved
```{r}
# Create a column for human vs non-human
combined_metadata_norm$human_status <- factor(ifelse(
  combined_metadata_norm$Species == "Homo sapiens",
  "Human",
  "Non-human"
))
```

```{r}
# Define batch variable
batch_hu <- combined_metadata_norm$human_status

# Set sample names to batches
names(batch_hu) <- combined_metadata_norm$Sample
```

```{r}
# Define biological covariates to preserve
# preserve CancerTypeFinal
design <- model.matrix(~CancerTypeFinal, data = combined_metadata_norm)

```

```{r}
# Batch correct
bc_expr_norm <- ComBat(dat = combined_expr_norm, batch = batch_hu, mod = design)

```

### PCA of above batch corrected data

```{r}
pca_after_bc <- visualize_pca(
  exp = bc_expr_norm,
  met = combined_metadata_norm,
  title = "Human vs Non-human normalized + batch corrected + CT preserved",
  group_col = "pca_group",
  colors = pca_colors,
  legend_title = "Species"
)
```


```{r}
library(Polychrome)
data("glasbey")

# Function to view pca colored by CancerSystem
visualize_pca_cs <- function(exp, met, title, color_by = "CancerSystem", custom_colors = NULL){
  # Set Genes as row names
  all_exp_filtered_rem <- exp %>% column_to_rownames("Genes")
  
  # Transpose the data (genes in columns, samples in rows)
  exp_transposed <- t(all_exp_filtered_rem)
  
  # Remove zero-variance genes
  exp_subset <- exp_transposed[, which(apply(exp_transposed, 2, var) != 0), drop = FALSE]
  
  # Perform PCA
  df_pca <- prcomp(exp_subset, scale = TRUE, center = TRUE)
  
  # Prepare PCA dataframe
  df <- data.frame(
    PC1 = df_pca$x[,1],
    PC2 = df_pca$x[,2],
    ColorGroup = met[[color_by]],
    Species = if ("Species" %in% colnames(met)) met$Species else NA
  )
  
  # If no custom colors, use viridis
  if (is.null(custom_colors)) {
    groups <- unique(df$ColorGroup)
    n_colors <- length(groups)
    custom_colors <- glasbey[1:n_colors]
    names(custom_colors) <- groups
  }
  
  # Base ggplot setup
  pca_plot <- ggplot(df, aes(x = PC1, y = PC2)) +
    labs(x = "PC1", y = "PC2", color = "Cancer System") +
    theme_minimal() + # white background
    theme(
      axis.line = element_line(colour = "black"), # bold axis lines
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA)
    )
  
  pca_plot <- pca_plot + 
    geom_point(aes(color = ColorGroup), size = 2, alpha = 0.6)

  # Apply color scale here (always after geom_point)
  pca_plot <- pca_plot + scale_color_manual(values = custom_colors)
  
  # Save PCA plot as pdf
  ggsave(
    filename = paste0(title, ".pdf"),
    plot = pca_plot,
    path = here(plot_folder),
    width = 10, height = 8,
    #limitsize = FALSE,
    device = cairo_pdf
  )
  
  return(pca_plot)
}

```

```{r}
# Check unique Cancer Systems
unique(combined_metadata_norm$CancerSystem)
```

```{r}
# Rename 'Head_and_Neck' to 'Head and Neck'
combined_metadata_norm <- combined_metadata_norm %>%
  mutate(CancerSystem = recode(CancerSystem,
    "Head_and_Neck" = "Head and Neck"
  ))

# Check
unique(combined_metadata_norm$CancerSystem)
```


```{r}
cs_colors <- c(
  "Breast" = "#4E79A7",
  "Digestive_Gastrointestinal" = "#F28E2B",
  "Endocrine_Neuroendocrine" = "#E15759",
  "Eye" = "#76B7B2",
  "Genitourinary" = "#59A14F",
  "Gynecologic" = "#EDC948",
  "Head and Neck" = "#B07AA1",
  "Hematologic_Blood" = "#FF9DA7",
  "Musculoskeletal" = "#9C755F",
  "Neurologic" = "#BAB0AC",
  "Respiratory_Thoracic" = "#1F77B4",
  "Skin" = "#AEC7E8",
  "Unspecified" = "#17BECF"
)

# Check
cs_colors

```

```{r}
# Make bc_expr_norm a dataframe 
bc_expr_norm <- as.data.frame(bc_expr_norm)

head(bc_expr_norm)
```

```{r}
# Edit metadata to label CancerSystem without underscores, replace with '/'
met_pca <- combined_metadata_norm
met_pca$CancerSystem_plot <- gsub("_", "/", met_pca$CancerSystem)


# CancerSystem pca normalized
cs_norm_pca <- visualize_pca_cs(bc_expr_norm %>% rownames_to_column("Genes"), met_pca, "Human vs Non-human normalized + batch corrected + CT preserved CancerSystem colored", color_by = "CancerSystem_plot", custom_colors = setNames(cs_colors, gsub("_", "/", names(cs_colors))))

```

### Output PCAs as a figure in a 1x1 grid 

```{r}
library(patchwork)
```

```{r}
# Create grid with PCA after normalization no duplicates colored by species and colored by cancer system
pca_fig <- (pca_after_bc + cs_norm_pca) + plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 22),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 16),
        legend.key.size = unit(0.6, "cm")) &
  guides(
    color = guide_legend(override.aes = list(size = 3))
  )

pca_fig
```

```{r}
# Save figure as pdf
ggsave(here(plot_folder, "AtlasWithHumansPCA.pdf"), pca_fig, width = 25, height = 10, device = cairo_pdf)

```


```{r}
# Check
all(colnames(bc_expr_norm) == combined_metadata_norm$Sample)
```

## Output treehouse + mammals metadata and exp data

```{r}
# Save simplified metadata
saveRDS(combined_metadata_norm, file = here("output_data/non_mouse_data/updated_data/species+human/treehouse_mammals_no_dup_metadata_5-2-26.rds"))

write_tsv(combined_metadata_norm, file = here("output_data/non_mouse_data/updated_data/species+human/treehouse_mammals_no_dup_metadata_5-2-26.tsv"))

# Save exp data
saveRDS(bc_expr_norm %>% rownames_to_column("Genes"), file = here("output_data/non_mouse_data/updated_data/species+human/treehouse_mammals_no_dup_bc_exp_tmm_logcpm.rds"))

write_tsv(bc_expr_norm %>% rownames_to_column("Genes"), file = here("output_data/non_mouse_data/updated_data/species+human/treehouse_mammals_no_dup_bc_exp_tmm_logcpm.tsv"))

```

